---
title: "Disturbance–response analysis II — wavelet synchrony (coherence, cross-power, phase → apparent lag, amplitude)"
subtitle: "Robust signal-processing pass: Morlet wavelet at the P≈500 band, pre vs post — the native wavelet view, read against analysis I's moments + peaks"
author: "nic (nic@cascadeclimate.org)"
date: "`r Sys.Date()`"
# Knit → render in place, then copy the self-contained HTML to <repo-root>/docs for GitHub Pages.
knit: |
  (function(input, encoding) {
    out  <- rmarkdown::render(input, encoding = encoding)
    docs <- file.path(dirname(normalizePath(input)), "..", "docs")
    dir.create(docs, showWarnings = FALSE, recursive = TRUE)
    file.copy(out, docs, overwrite = TRUE)
    out
  })
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    fig_width: 10
    fig_caption: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
if (!requireNamespace("WaveletComp", quietly = TRUE))                  # WaveletComp is required
  stop("analysis_02 needs WaveletComp — install once with  install.packages('WaveletComp')")
suppressMessages({ library(dplyr); library(tidyr); library(readr); library(ggplot2)
                   library(knitr); library(patchwork); library(WaveletComp) })
# Deterministic — no RNG / significance surrogates, so re-knits reproduce every number exactly.
theme_set(theme_bw(base_size = 11) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom"))
P  <- 500              # generator period (steps); see CLAUDE.md. The band we read synchrony at.
EF <- round(sqrt(2) * P)  # Morlet e-folding (~707) = time support at period P. Serves as both the
                          # COI margin from the series ends and the buffer from each regime boundary
                          # (so a settled-window average never straddles the disturbance/transition).
fig_n <- local({ i <- 0L; function(x) { i <<- i + 1L; paste0("Figure ", i, ". ", x) } })
```

Analysis I described synchrony with **moments + peak spacings** over frozen `pre`/`post`
windows — the simplest thing that works, but peak-finding frays on noisy/irregular signals
and its lag is a coarse few-sample offset. This pass takes the **native wavelet view**: the
Morlet wavelet (via **WaveletComp**) resolves synchrony jointly in **time and frequency**, so
we can isolate the generator's `P≈500` band and watch coupling evolve *through* the
disturbance, not just at the settled ends. Deliberately **different metrics** from analysis I
— the point is to compare what each approach sees (there is no attempt to reproduce the
peak-based numbers).

Four band-localized quantities, all read at `P≈500` (the driver's period):

- **Wavelet coherence** `∈[0,1]` — normalized (smoothed) coupling strength.
- **Cross-wavelet power** — strength of *shared* oscillation at the band (un-normalized).
- **Phase difference φ** (radians) — `Angle = phase.climate − phase.response`, the raw synchrony
  phase; **climate leads ⇒ positive φ**. Shown in the tables alongside its step-unit form:
- **Apparent lag** = `φ · P / (2π)` steps — φ rescaled to the interpretable "how many steps
  behind" (the figures use this; the tables carry both).
- **Amplitude ratio** — physical band amplitude `response / climate` (see the `×sd` note in
  `band_trace`: WaveletComp standardizes each channel, so the raw amplitude ratio must be
  rescaled back to physical units).

*Two caveats carried throughout.* **(1) COI/edge:** at period `P` the Morlet spans
`~±707` steps, so band metrics are only trustworthy `≥707` steps from a series end *and* from
a regime boundary — settled averages use those buffered windows, and trace edges are blanked.
**(2) Uncertainty:** the band series is heavily autocorrelated, so error bars use a
**block SE** — one estimate per `P`-long cycle, `SE = dispersion/√n_blocks` (a *circular*
dispersion for the phase, since linear sd on wrapped angles is invalid). The COI buffer leaves
only **~2 cycles per settled window** (reported as `n_blocks`), so these SEs carry ~1 degree of
freedom: read them as **coarse guides** to whether an effect clears its noise, not precise
intervals. Shown next to each pre→post change so effect size is read against its uncertainty.

## Helpers

```{r helpers}
# ---- small numerics -------------------------------------------------------------------------
smooth_ma <- function(x, w = 51) as.numeric(stats::filter(x, rep(1 / w, w)))  # centered MA; NA at ±(w-1)/2 ends
CMEAN     <- function(a) Arg(mean(exp(1i * a)))          # circular mean of angles (phase never averaged linearly)
band_rows <- function(wc) which(abs(wc$Period - P) <= 0.05 * P)  # the few scale-rows within ±5% of P

# wc.image styling is identical across all three scenario spectra — only the object + title vary.
# color.key = "interval": linear 0→1 colour scale. ("quantile" warps it — coherence saturates
# near 1, so quantile breaks bunch the legend into a 0/0.9/1 mess.)
plot_wc <- function(wc, main)
  wc.image(wc, which.image = "wc", plot.arrow = TRUE, plot.contour = FALSE,
           color.key = "interval", timelab = "time", periodlab = "period",
           legend.params = list(lab = "wavelet coherence"), main = main)

# ---- WaveletComp wrapper --------------------------------------------------------------------
# Analyzed on climate vs log(response) (log-linear generator).
# loess.span = 0: series are trend-free ⇒ standardization is exactly /sd(x), making the ×sd
#   amplitude rescale in band_trace exact.
# my.pair ⇒ Angle = phase.climate − phase.log_response, so climate-leads is +.
# make.pval = FALSE: significance surrogates skipped — a valid 5% contour needs ≥100 (~30-50 min)
#   and coherence saturates here, so it would mislead; uncertainty = COI + block SEs instead.
# lower/upperPeriod bracket P by ±1.5 octaves so the coupling ridge sits mid-axis (upper ≈ the
#   series-length/COI limit) and the scale grid anchors ~onto P.
wco <- function(df)
  analyze.coherency(as.data.frame(df[, c("climate", "log_response")]),
                    my.pair = c("climate", "log_response"), loess.span = 0,
                    dt = 1, dj = 1/20, lowerPeriod = P / 2^1.5, upperPeriod = P * 2^1.5,
                    make.pval = FALSE, verbose = FALSE)

# ---- collapse the wc object to one time-resolved series at the P band ------------------------
band_trace <- function(wc, df) {
  bi   <- band_rows(wc)
  sd_c <- sd(df$climate); sd_r <- sd(df$log_response)      # the constants WaveletComp divided out (loess.span=0)
  ang  <- apply(wc$Angle[bi, , drop = FALSE], 2, CMEAN)    # per-time band phase difference
  tibble(time      = df$time,
         phase     = df$phase,
         coherence = colMeans(wc$Coherence[bi, , drop = FALSE]),
         xpower    = colMeans(wc$Power.xy[bi, , drop = FALSE]),                 # shared-oscillation power at band
         amp_ratio = colMeans(wc$Ampl.y[bi, , drop = FALSE]) /
                     colMeans(wc$Ampl.x[bi, , drop = FALSE]) * (sd_r / sd_c),   # ×sd: undo standardization → log-space band-amp ratio ≈ β
         angle     = ang,
         lag       = ang / (2 * pi) * P,                                        # phase → steps
         coi_ok    = df$time >= EF & df$time <= nrow(df) - EF)                  # trustworthy (outside COI) mask
}

# ---- settled pre/post summary with block SEs ------------------------------------------------
# One estimate per P-long cycle in a COI-buffered window, then value ± dispersion/√n_blocks.
# Valid under autocorrelation (a per-step SE would be far too narrow). `circ` uses a circular
# dispersion — linear sd on wrapped angles is invalid (phases straddling ±π look far apart).
# The COI buffer leaves only ~2 cycles per window, so n is tiny — SEs are coarse (see intro).
block_stat <- function(t, v, circ = FALSE) {
  b  <- (t - min(t)) %/% P
  bv <- as.numeric(tapply(seq_along(t), b, function(i) if (circ) CMEAN(v[i]) else mean(v[i])))
  n  <- length(bv)
  if (circ) {
    R    <- Mod(mean(exp(1i * bv)))                       # resultant length of block phases (1 = locked, 0 = scattered)
    disp <- if (R < 1) sqrt(-2 * log(R)) else 0           # circular sd
  } else disp <- sd(bv)
  list(value = if (circ) CMEAN(bv) else mean(bv),
       se    = if (n > 1) disp / sqrt(n) else NA_real_, n = n)
}

# Regime boundaries come from the `phase` column itself — one source of truth, so the settled
# windows can't silently desync from the data (the disturbance step and transition span differ
# by scenario; Return's transition runs to 2500, the others to 2250).
wave_summary <- function(tr) {
  td     <- tr$time[tr$phase == "disturbance"]            # the single disturbance step
  tr_end <- max(tr$time[tr$phase == "transition"])        # last step of the transition span
  N      <- max(tr$time)
  wins <- list(pre  = tr$time >= EF          & tr$time <= td - EF,      # ≥ EF from series start AND from disturbance
               post = tr$time >= tr_end + EF & tr$time <= N  - EF)      # ≥ EF from transition end AND from series end
  one <- function(sel) {
    t  <- tr$time[sel]
    co <- block_stat(t, tr$coherence[sel]); xp <- block_stat(t, tr$xpower[sel])
    ar <- block_stat(t, tr$amp_ratio[sel]); an <- block_stat(t, tr$angle[sel], circ = TRUE)
    tibble(n_blocks  = co$n,                                            # cycles the SEs rest on (expect ~2)
           coherence = co$value, coherence_se = co$se,
           xpower    = xp$value, xpower_se    = xp$se,
           amp_ratio = ar$value, amp_ratio_se = ar$se,
           phi       = an$value,                 phi_se = an$se,          # phase difference φ, radians (circular SE)
           lag       = an$value / (2*pi) * P, lag_se = an$se / (2*pi) * P)  # same φ in step units
  }
  bind_rows(pre = one(wins$pre), post = one(wins$post), .id = "phase") %>%
    mutate(phase = factor(phase, c("pre", "post")))
}

# ---- model-based residual (whole series, log space) -----------------------------------------
# Log-linear generator ⇒ predict log(response): predicted = gain · climate(t − lag), residual =
# log(response) − predicted. lag & gain are fit on the `pre` window only (cross-correlation peak +
# no-intercept OLS), NOT read from the generator, so the residual sits at ~0 through pre and
# departs exactly where coupling changes.
fit_pre <- function(df) {
  pre <- df[df$phase == "pre", ]; cl <- pre$climate; rs <- pre$log_response; n <- length(rs)
  Ls  <- 0:round(0.6 * P)                                              # physically plausible lags only (< ~0.6 period)
  cr  <- vapply(Ls, function(L) cor(rs[(L + 1):n], cl[seq_len(n - L)]), numeric(1))  # cor(log(response)(t), climate(t−L))
  L   <- Ls[which.max(cr)]
  x   <- cl[seq_len(n - L)]; y <- rs[(L + 1):n]
  list(lag = L, gain = sum(x * y) / sum(x * x))                        # OLS slope, no intercept (channels ~zero-mean)
}

resid_trace <- function(df) {
  fp <- fit_pre(df); L <- fp$lag; N <- nrow(df)
  pred <- rep(NA_real_, N); if (L < N) pred[(L + 1):N] <- fp$gain * df$climate[seq_len(N - L)]
  resid <- df$log_response - pred
  out <- tibble(time = df$time, phase = df$phase,
                resid  = resid,
                dresid = c(NA, diff(smooth_ma(resid, 51))))            # smooth THEN diff so d/dt isn't pure noise
  attr(out, "lag") <- L; attr(out, "gain") <- fp$gain; out
}

# ---- shared plotting furniture (band shading matches analysis I) ----------------------------
phase_fill <- c(pre = "#9ecae1", disturbance = "grey45", transition = "#fcbba1", post = "#fdae6b")
bar_fill   <- phase_fill[c("pre", "post")]                              # the bar plots reuse the pre/post band colours
phase_bands <- function(ph) ph %>% filter(phase != "disturbance") %>%   # disturbance is one step → drawn as a line
  group_by(phase) %>% summarise(xmin = min(time), xmax = max(time), .groups = "drop")

# shared band shading: rect per phase + dashed disturbance line + fill scale.
phase_band_layers <- function(d) {
  bands <- phase_bands(d); td <- d$time[d$phase == "disturbance"]
  list(geom_rect(data = bands, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = phase),
                 alpha = 0.30, inherit.aes = FALSE),
       geom_vline(xintercept = td, linetype = 2, colour = "grey30"),
       scale_fill_manual(values = phase_fill, guide = "none"))
}

# whole-series residual + its smoothed derivative on a secondary axis, phases shaded AND labelled.
plot_resid <- function(df, title) {
  rt <- resid_trace(df); L <- attr(rt, "lag"); g <- attr(rt, "gain")
  bands <- phase_bands(df)                                             # kept for the phase labels below
  ytop <- max(abs(rt$resid), na.rm = TRUE)
  sc   <- ytop / max(abs(rt$dresid), na.rm = TRUE)                     # scale d/dt onto the residual axis (2nd axis un-scales it)
  ggplot() +
    phase_band_layers(df) +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
    geom_line(data = rt, aes(time, dresid * sc), colour = "#d94801", linewidth = 0.3, na.rm = TRUE) +
    geom_line(data = rt, aes(time, resid),       colour = "#238b45", linewidth = 0.4, na.rm = TRUE) +
    geom_text(data = mutate(bands, x = (xmin + xmax) / 2),
              aes(x = x, y = ytop, label = phase), vjust = 1, size = 3, colour = "grey25") +
    scale_y_continuous(sec.axis = sec_axis(~ . / sc, name = "d(residual)/dt  (smoothed, orange)")) +
    labs(title = title, x = "time", y = "residual = log(response) − predicted  (green)",
         subtitle = sprintf("pre-fit predictor:  log(response)(t) ≈ %.2f · climate(t − %d)", g, L))
}

# time-resolved band metrics; COI-contaminated edges blanked; phases shaded.
metric_labs <- c(coherence = "coherence", xpower = "cross power", amp_ratio = "amp ratio", lag = "app. lag")
plot_band_trace <- function(tr, title) {
  m <- names(metric_labs)                                              # the four plotted metrics (single source)
  long <- tr %>%
    mutate(across(all_of(m), ~ ifelse(coi_ok, ., NA_real_))) %>%
    select(time, all_of(m)) %>%
    pivot_longer(-time, names_to = "metric") %>%
    mutate(metric = factor(metric, m))
  ggplot() +
    phase_band_layers(tr) +
    geom_line(data = long, aes(time, value), linewidth = 0.35, na.rm = TRUE) +
    facet_wrap(~ metric, scales = "free_y", ncol = 1, strip.position = "left",
               labeller = as_labeller(metric_labs)) +
    labs(title = title, x = "time", y = NULL)
}

# reshape a summary tibble to long (id keys × metric × value × se), driven by metric_labs +
# the regular `_se` naming — the one place the value↔se pairing lives. Used by the bars and summary.
metrics_long <- function(sm, keys) {
  m <- names(metric_labs)
  val <- sm %>% select(all_of(c(keys, m))) %>%
    pivot_longer(all_of(m), names_to = "metric", values_to = "value")
  se  <- sm %>% select(all_of(c(keys, paste0(m, "_se")))) %>%
    rename_with(~ sub("_se$", "", .x), all_of(paste0(m, "_se"))) %>%
    pivot_longer(all_of(m), names_to = "metric", values_to = "se")
  left_join(val, se, by = c(keys, "metric")) %>% mutate(metric = factor(metric, m))
}

# pre/post bars per metric, ±1 SE (block SE) so effect size is read against uncertainty.
plot_wave_bars <- function(sm, title) {
  dat <- metrics_long(sm, "phase")
  ggplot(dat, aes(phase, value, fill = phase)) +
    geom_col(width = 0.7) +
    geom_errorbar(aes(ymin = value - se, ymax = value + se), width = 0.25, colour = "grey25", na.rm = TRUE) +
    facet_wrap(~ metric, scales = "free_y", nrow = 1, labeller = as_labeller(metric_labs)) +
    scale_fill_manual(values = bar_fill, guide = "none") +
    labs(title = title, x = NULL, y = NULL)
}

# the settled pre/post table — same columns every scenario (adds n_blocks + phi, no _se).
emit_table <- function(sm, caption)
  cat(kable(sm %>% select(phase, n_blocks, coherence, xpower, amp_ratio, phi, lag),
            digits = 2, caption = caption), sep = "\n")
```

```{r load}
# response is analyzed in log space (log-linear generator) → add log_response once at load.
d_shift  <- read_csv("synthetic_shift.csv",  show_col_types = FALSE) %>% mutate(log_response = log(response))
d_desync <- read_csv("synthetic_desync.csv", show_col_types = FALSE) %>% mutate(log_response = log(response))
d_return <- read_csv("synthetic_return.csv", show_col_types = FALSE) %>% mutate(log_response = log(response))
```

```{r derive}
# One coherency transform per dataset → reused by every image, trace and table below.
wc_shift  <- wco(d_shift);  tr_shift  <- band_trace(wc_shift,  d_shift);  sm_shift  <- wave_summary(tr_shift)
wc_desync <- wco(d_desync); tr_desync <- band_trace(wc_desync, d_desync); sm_desync <- wave_summary(tr_desync)
wc_return <- wco(d_return); tr_return <- band_trace(wc_return, d_return); sm_return <- wave_summary(tr_return)
```

---

# Shift

Expectation: coupling *survives* the disturbance but the regime changes — response log-amplitude
halves (band amp ratio `1.0→0.5`, = β) and lag shortens (`50→10`). Wavelet reading: coherence and
cross-power stay high; **apparent lag drops ~50→10** and **amplitude ratio drops ~1.0→0.5**.

```{r shift-resid, fig.height=3.4, fig.cap=fig_n("Shift — model residual (log(response) − pre-fit prediction, green) over the whole series with its smoothed derivative (orange, right axis); phases shaded and labelled, dashed = disturbance. Flat ~0 through pre by construction; a new nonzero regime after the disturbance.")}
plot_resid(d_shift, "Shift — model residual & its derivative")
```

```{r shift-wc, fig.height=4.6, fig.cap=fig_n("Shift — wavelet coherence spectrum (time × period). Arrows = phase (right = in-phase, down = climate leads); the pale COI parabola marks edge-unreliable regions. Coupling holds across the disturbance at the P≈500 band.")}
plot_wc(wc_shift, "Shift — wavelet coherence")
```

```{r shift-trace, fig.height=5.2, fig.cap=fig_n("Shift — band (P≈500) metrics through time; COI edges blanked, phases shaded. Coherence & cross-power hold; apparent lag steps 50→10 and amplitude ratio 1.0→0.5 across the transition.")}
plot_band_trace(tr_shift, "Shift — synchrony at the P≈500 band, through time")
```

```{r shift-table, results='asis'}
emit_table(sm_shift, "Shift — band synchrony, settled pre vs post (n_blocks = cycles the SEs rest on; SEs as error bars)")
```

```{r shift-bars, fig.height=3, fig.cap=fig_n("Shift — pre vs post band metrics, ±1 SE (block SE). The amp-ratio halving and lag shortening dwarf their error bars; coherence/cross-power barely move — a coherent regime shift.")}
plot_wave_bars(sm_shift, "Shift — band metrics, pre vs post (±1 SE)")
```

Read-off: amplitude ratio ≈ **1.0 → 0.5** and apparent lag ≈ **50 → 10** while coherence and
cross-power stay high — the wavelet agrees with analysis I's "new regime, still coupled," and
pins the lag change far more sharply than peak offsets could.

---

# Desync

Expectation: `post` response is stationary noise matched to regime-n *variance* — so its
energy leaves the `P` band and spreads across all scales. Wavelet reading: **cross-power and
amplitude ratio at the band collapse**; note whether the *normalized* coherence follows.

```{r desync-resid, fig.height=3.4, fig.cap=fig_n("Desync — model residual (green) + derivative (orange, right axis), whole series; phases shaded, dashed = disturbance. Residual is ~0 noise in pre, then becomes large and erratic as the response decouples into noise.")}
plot_resid(d_desync, "Desync — model residual & its derivative")
```

```{r desync-wc, fig.height=4.6, fig.cap=fig_n("Desync — wavelet coherence spectrum (arrows = phase; pale COI parabola = edge-unreliable). Watch the P≈500 band after the disturbance: the smoothed coherence stays high (WaveletComp's smoothing is too narrow relative to scale, so coherence saturates toward 1 even over noise), so coherence alone does not flag the decoupling — cross-power does.")}
plot_wc(wc_desync, "Desync — wavelet coherence")
```

```{r desync-trace, fig.height=5.2, fig.cap=fig_n("Desync — band metrics through time. Cross-power and amplitude ratio crash toward 0 after the disturbance while coherence stays high — the decoupling shows in POWER, not coherence.")}
plot_band_trace(tr_desync, "Desync — synchrony at the P≈500 band, through time")
```

```{r desync-table, results='asis'}
emit_table(sm_desync, "Desync — band synchrony, settled pre vs post (n_blocks = cycles the SEs rest on)")
```

```{r desync-bars, fig.height=3, fig.cap=fig_n("Desync — pre vs post band metrics, ±1 SE. Cross-power and amplitude ratio collapse (well beyond their SEs); coherence stays high and the apparent lag SE balloons (phase no longer stable) — its residual value is meaningless.")}
plot_wave_bars(sm_desync, "Desync — band metrics, pre vs post (±1 SE)")
```

Read-off: the key limitation of the **native** view — WaveletComp's *normalized, smoothed*
coherence stays near 1 even though the response is noise, and a phase/lag is still numerically
returned (with a huge SE). What actually registers the decoupling is the **un-normalized
cross-power and the band amplitude ratio collapsing to ≈0**: the watershed no longer
oscillates at the climate period. (Analysis I caught the same event via peak-spacing
irregularity — a different fingerprint of the same decoupling.)

---

# Return

Expectation: the excursion peaks then recovers to **regime n**, so `pre` and `post` are the
same regime — every band metric should match end-to-end.

```{r return-resid, fig.height=3.4, fig.cap=fig_n("Return — model residual (green) + derivative (orange, right axis), whole series; phases shaded, dashed = disturbance. Residual departs during the excursion (derivative spikes at onset and recovery) then settles back to ~0.")}
plot_resid(d_return, "Return — model residual & its derivative")
```

```{r return-wc, fig.height=4.6, fig.cap=fig_n("Return — wavelet coherence spectrum. Coupling at the P≈500 band is maintained end-to-end; the transient excursion perturbs phase/power only within the transition window.")}
plot_wc(wc_return, "Return — wavelet coherence")
```

```{r return-trace, fig.height=5.2, fig.cap=fig_n("Return — band metrics through time. A transient wobble during the excursion, then all four metrics return to their pre values by the settled post window.")}
plot_band_trace(tr_return, "Return — synchrony at the P≈500 band, through time")
```

```{r return-table, results='asis'}
emit_table(sm_return, "Return — band synchrony, settled pre vs post (n_blocks = cycles the SEs rest on)")
```

```{r return-bars, fig.height=3, fig.cap=fig_n("Return — pre vs post band metrics, ±1 SE. Every metric matches within error — no lasting signature.")}
plot_wave_bars(sm_return, "Return — band metrics, pre vs post (±1 SE)")
```

Read-off: pre ≈ post on all four band metrics — the disturbance leaves **no lasting wavelet
signature**, matching analysis I's null result.

---

# Transition frequency — the Doppler chirp

The settled band metrics above read synchrony only at `P=500 ±5%` and buffer the transition out,
so they miss what happens to *frequency* during the change. A shortening lag **is** a frequency
shift: with `response(t) = f(climate(t − τ(t)))`, the instantaneous frequency is
`f_out = f_in·(1 − dτ/dt)` — while τ shrinks the response runs faster, then returns to `P`. Both
regimes share `P`, so there is **no permanent** shift (post frequency returns exactly to `P`), but
the transition carries a **real** transient excursion — generator signal, not a numerical artifact.
A ~16% excursion (period → ~430) sits *outside* the ±5% band, which is why the settled metrics
correctly ignore it and a dedicated frequency read is needed to see it.

Measured two ways and compared. **Measured:** instantaneous response period = the ridge (per-time
argmax over period) of its wavelet amplitude. **Predicted, from the data:** recover the lag at the
pre end, the transition midpoint, and the post end by cross-correlation, and the transition span
`W` from the `phase` column; `τ(t)` piecewise-linear through those anchors gives period
`P/(1 − dτ/dt)`, peak `|Δf/f| ≈ Δlag/W`.

*Caveat: at period `P` the Morlet spans ~±707 steps (the e-folding `EF`) — wider than the 250-step
transition — so the measured ridge excursion is time-smeared and attenuated relative to the clean
prediction. The prediction is the kinematic bound; the measurement confirming its sign and timing
is the point, not an exact magnitude match.*

```{r chirp-helpers}
# Instantaneous response period = ridge (per-time argmax over the period axis) of the response
# wavelet amplitude. Dips below P where the response oscillation compresses.
inst_period <- function(wc, df)
  tibble(time = df$time, phase = df$phase,
         period = wc$Period[apply(wc$Ampl.y, 2, which.max)],
         coi_ok = df$time >= EF & df$time <= nrow(df) - EF)

# Lag L maximizing cor(climate(t−L), log_response(t)) over a row mask (same scan as fit_pre).
win_lag <- function(df, sel) {
  cl <- df$climate; rs <- df$log_response; i0 <- which(sel); Ls <- 0:round(0.6 * P)
  cr <- vapply(Ls, function(L) { i <- i0[i0 > L]; cor(rs[i], cl[i - L]) }, numeric(1))
  Ls[which.max(cr)]
}

# Predicted transition period from data-recovered endpoint lags. response(t)=climate(t−τ(t)) ⇒
# f_out=f_in·(1−τ′) ⇒ period=P/(1−τ′). Monotonic transition ⇒ constant τ′=(lag_post−lag_pre)/W.
# Only meaningful when endpoints differ (Shift); Return's endpoints match and Desync's post-lag is
# noise, so those are shown measured-only.
chirp_predict <- function(df) {
  td <- df$time[df$phase == "disturbance"]
  tr <- df$time[df$phase == "transition"]; t0 <- min(tr); t1 <- max(tr); W <- length(tr)
  lag_pre  <- win_lag(df, df$phase == "pre"  & df$time >= EF & df$time <= td - EF)
  lag_post <- win_lag(df, df$phase == "post" & df$time >= t1 + EF)
  dtau <- (lag_post - lag_pre) / W
  list(pred = tibble(time = c(t0, t1), period = P / (1 - dtau)),   # horizontal predicted level
       lag_pre = lag_pre, lag_post = lag_post, W = W, frac = abs(dtau))
}

# measured ridge (green) + optional data-derived Doppler prediction (dashed); y clipped to ±30% P.
plot_chirp <- function(wc, df, title, predict = TRUE) {
  ip <- inst_period(wc, df) %>% mutate(period = ifelse(coi_ok, period, NA_real_))
  g <- ggplot() + phase_band_layers(df) +
    geom_hline(yintercept = P, colour = "grey55", linewidth = 0.3) +
    geom_line(data = ip, aes(time, period), colour = "#238b45", linewidth = 0.4, na.rm = TRUE) +
    coord_cartesian(ylim = c(0.7 * P, 1.3 * P)) +
    labs(title = title, x = "time", y = "instantaneous period (response amplitude ridge)")
  if (predict) {
    cp <- chirp_predict(df)
    g <- g + geom_line(data = cp$pred, aes(time, period), colour = "#d94801",
                       linewidth = 0.6, linetype = 2, na.rm = TRUE) +
      labs(subtitle = sprintf(
        "data-derived lags %d→%d, W=%d ⇒ predicted |Δf/f| ≈ %.0f%%",
        cp$lag_pre, cp$lag_post, cp$W, 100 * cp$frac))
  }
  g
}
```

```{r chirp-shift, fig.height=3.4, fig.cap=fig_n("Shift — instantaneous response period (green ridge) vs the data-derived Doppler prediction (dashed orange). One excursion: the response speeds up (period dips below P) as the lag shortens 50→10, then settles at P.")}
plot_chirp(wc_shift, d_shift, "Shift — one chirp (lag 50→10)")
```

```{r chirp-return, fig.height=3.4, fig.cap=fig_n("Return — two excursions: the response speeds up toward the peak then slows on the way back (period dips below then rises above P), re-settling at P. Endpoints are the same regime (lag ~50 both), so this two-limb signature lives entirely in the transition — invisible to any endpoint-based read; shown measured-only, and attenuated by Morlet smearing over the 500-step excursion.")}
plot_chirp(wc_return, d_return, "Return — two chirps (lag 50→10→50)", predict = FALSE)
```

```{r chirp-desync, fig.height=3.4, fig.cap=fig_n("Desync — no chirp: the lag never changes (w=0). Instead the response's P-band power collapses (cross-power trace above), so past the disturbance there is no coherent ridge — the argmax wanders. Prediction overlay omitted (recovered post-lag is meaningless on noise).")}
plot_chirp(wc_desync, d_desync, "Desync — no chirp (power leaves the band)", predict = FALSE)
```

Together: **Shift 1** excursion, **Return 2**, **Desync 0** — the frequency-domain signature of
each scenario's lag trajectory, invisible to the fixed-band metrics and to a linear-scale read.

---

# Cross-dataset summary

Post-minus-pre change in the four band metrics, per scenario, each Δ carrying its propagated
SE (`√(se_pre² + se_post²)`) so the change is read against its uncertainty.

```{r summary}
sm_all <- bind_rows(Shift = sm_shift, Desync = sm_desync, Return = sm_return, .id = "dataset")

# Long form (metric × value × se) via the shared reshape, then pre→post delta + propagated SE.
delta <- metrics_long(sm_all, c("dataset", "phase")) %>%
  pivot_wider(names_from = phase, values_from = c(value, se)) %>%
  mutate(delta    = value_post - value_pre,
         delta_se = sqrt(se_pre^2 + se_post^2),
         dataset  = factor(dataset, c("Shift", "Desync", "Return")))
```

```{r summary-table, results='asis'}
cat(kable(delta %>% transmute(dataset, metric = metric_labs[as.character(metric)],
                              pre = value_pre, post = value_post, delta, delta_se),
          digits = 2, caption = "pre → post change in band metrics, all datasets (delta_se = propagated ±1 SE)"), sep = "\n")
```

```{r summary-fig, fig.height=4, fig.cap=fig_n("Post − pre change per band metric, faceted by scenario, ±1 propagated SE. Shift = amp-ratio & lag drop (coupled regime shift); Desync = cross-power & amp-ratio collapse (decoupling, invisible to coherence); Return = all Δ ≈ 0.")}
ggplot(delta, aes(metric, delta, fill = metric)) +
  geom_col() +
  geom_errorbar(aes(ymin = delta - delta_se, ymax = delta + delta_se), width = 0.25, colour = "grey25", na.rm = TRUE) +
  geom_hline(yintercept = 0, colour = "grey50") +
  facet_wrap(~ dataset) +
  scale_x_discrete(labels = metric_labs) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(title = "Post − pre change by scenario (±1 SE)", x = NULL, y = "post − pre") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

**Takeaways — the native wavelet view, next to analysis I.** The Morlet band reading pins the
**Shift** cleanly: amplitude ratio and apparent lag move together (1.0→0.5, 50→10) while
coherence holds — the same "coupled regime shift" analysis I inferred, but with a sharp,
sub-step lag instead of a few-sample peak offset. **Return** is a flat null in both views. The
instructive disagreement is **Desync**: the wavelet's *normalized* coherence is **fooled**
(stays ≈1, and still hands back a phase/lag), so the decoupling reads only in the
**un-normalized cross-power and band amplitude** collapsing to ≈0 — where analysis I saw it as
peak-spacing irregularity. Two methods, two fingerprints of the same event: the lesson
is that *coherence alone is not enough* — you need a power/amplitude channel to see a driver
the response has stopped tracking (a significance test wouldn't rescue it: the saturated
coherence would itself read as "significant").
```
