---
title: "Real-data disturbance–response IV — wavelet synchrony (coherence, cross-power, phase → apparent lag, amplitude)"
subtitle: "The Morlet-wavelet approach on real MacroSheds precip→discharge — how the native signal-processing view holds up off the idealised bench (cf. analysis II), read against analysis III's peaks"
author: "nic (nic@cascadeclimate.org)"
date: "`r Sys.Date()`"
# Knit → render in place, then copy the self-contained HTML to <repo-root>/docs for GitHub Pages.
knit: |
  (function(input, encoding) {
    out  <- rmarkdown::render(input, encoding = encoding)
    docs <- file.path(dirname(normalizePath(input)), "..", "docs")
    dir.create(docs, showWarnings = FALSE, recursive = TRUE)
    file.copy(out, docs, overwrite = TRUE)
    out
  })
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: darkly
    highlight: breezedark
    fig_width: 10
    fig_caption: true
    self_contained: true
    includes:
      before_body: ../navbar.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dev.args = list(bg = "white"))   # native (light) PNG canvas; white matting added via CSS
if (!requireNamespace("WaveletComp", quietly = TRUE))                  # WaveletComp is required
  stop("analysis_04 needs WaveletComp — install once with  install.packages('WaveletComp')")
suppressMessages({ library(dplyr); library(tidyr); library(readr); library(ggplot2)
                   library(knitr); library(patchwork); library(WaveletComp) })
source("../R/wavelet_kernel.R")   # shared red-noise cross-power significance helpers (edit once)
# Deterministic — no RNG / significance surrogates, so re-knits reproduce every number exactly.
# (The significance MASK is analytic AR(1) red noise — Torrence & Compo 1998 — so it too is
#  deterministic: no make.pval / surrogates, reproducible from the tracked CSVs.)
theme_set(theme_bw(base_size = 11) + theme(legend.position = "bottom"))
P  <- 365              # seasonal (annual) period in days — the band we read synchrony at.
EF <- round(sqrt(2) * P)  # Morlet e-folding (~516 d) = time support at period P. Serves as both the
                          # COI margin from the series ends and the buffer from the disturbance
                          # (so a settled-window average never straddles the event).
# Phase determinacy. A phase difference is only defined where the DRIVER actually oscillates at the
# band. Precip is nearly aseasonal, so at P≈365 the driver amplitude is often ~noise and φ wanders
# across ±π (the ±P/2 "lag" sawtooth is that wrap, not a 6-month delay). Two guards, one meaning:
PHASE_MIN_AMP <- 0.5   # blank φ/lag in the TRACE where driver band-amplitude < this × its median (per time)
PHASE_MIN_R   <- 0.5   # suppress the SETTLED φ/lag where block-phase concentration R < this (scattered)
fig_n <- local({ i <- 0L; function(x) { i <<- i + 1L; paste0("Figure ", i, ". ", x) } })
```

```{css figure-matting, echo=FALSE}
/* White matting behind every figure: native light plots (incl. WaveletComp heatmaps) read cleanly on the darkly page. */
.main-container img,
div.figure img {
  background-color: #ffffff;
  padding: 10px;
  border-radius: 6px;
  box-sizing: border-box;
  max-width: 100%;
  box-shadow: 0 2px 10px rgba(0, 0, 0, 0.45);
}
```

This is **analysis II's method — Morlet wavelet coherence via WaveletComp — run on real
watersheds**, the wavelet counterpart to analysis III's peak/moments pass on the same five sites.
The study is a **methods comparison**: analysis II (`syn_data/analysis_02`) established, on a
clean synthetic system, what the native wavelet view sees that peaks cannot (coupling resolved
jointly in time *and* frequency; a sharp sub-step lag) — and its **characteristic failure**: the
*normalized, smoothed* coherence can be **fooled**, staying ≈1 even over noise, so the honest
signal lives in the **un-normalized cross-power and band amplitude**. Here we test whether those
strengths and that weakness reproduce on **less-deterministic real synchrony**: `climate` = daily
precipitation (undisturbed driver), `log(response)` = log daily discharge, read at the **annual
`P≈365 d` band**.

Four band-localized quantities, all read at `P≈365` (the seasonal band):

- **Wavelet coherence** `∈[0,1]` — normalized (smoothed) coupling strength.
- **Cross-wavelet power** — strength of *shared* oscillation at the band (un-normalized).
- **Phase difference φ** (radians) — `Angle = phase.climate − phase.response`; **climate leads ⇒
  positive φ**. Shown alongside its step-unit form:
- **Apparent lag** = `φ · P / (2π)` days — φ rescaled to "how many days behind." It is **gated to
  NA where the driver has no annual-band amplitude** (φ undefined there); a surviving value near
  `±½ yr` means *antiphase*, not a real six-month delay (see "Where the shared power lives").
- **Amplitude ratio** — band amplitude `response / climate` (rescaled back to physical units; see
  the `×sd` note in `band_trace`).

*Two caveats carried throughout.* **(1) COI/edge:** at period `P` the Morlet spans `~±516` days,
so band metrics are only trustworthy `≥516 d` from a series end *and* from the disturbance —
settled averages use those buffered windows, trace edges are blanked. **(2) Uncertainty:** the
band series is heavily autocorrelated, so error bars use a **block SE** — one estimate per `P`-long
(annual) cycle, `SE = dispersion/√n_blocks` (a *circular* dispersion for phase). Unlike the
synthetic windows (~2 cycles), the real windows hold **~15–28 years**, so these SEs are far better
resolved here.

**Provenance & framing.** Discharge is fully observed; **precip is partly modelled infill**
(negligible at hbef/fernow, ~13% at hjandrews). **`santa_barbara_MC06` is dropped** — its precip
driver was ~88% modelled infill (a wavelet there reads a modelled driver, not observed rain); it
stays in the EDA, where its coverage is now shown. And a
long-term trend sits *above* the analyzed range (upper period `730 d`), so `loess.span = 0` (no
detrending) is retained without trend-contaminating the seasonal read. Real precip is episodic,
not a clean sine, so — as analysis III already hinted — expect coherence to saturate and the
*power* channels to carry whatever real story there is.

## Helpers

```{r helpers}
# ---- small numerics -------------------------------------------------------------------------
smooth_ma <- function(x, w = 61) as.numeric(stats::filter(x, rep(1 / w, w)))  # centered MA; NA at ±(w-1)/2 ends
CMEAN     <- function(a) Arg(mean(exp(1i * a)))          # circular mean of angles (phase never averaged linearly)
band_rows <- function(wc) which(abs(wc$Period - P) <= 0.05 * P)  # the few scale-rows within ±5% of P

# wc.image styling is identical across all five site spectra — only the object + title vary.
# Horizontal dashed guides at a week / month / year. The period axis is log2, so a period p sits
# at y = log2(p); graphics.reset = FALSE keeps wc.image's coordinate system live to annotate over.
wc_marks <- c("1 wk" = 7, "1 mo" = 30, "3 mo" = 91, "1 yr" = 365)
plot_wc <- function(wc, df, main) {
  op <- par(bg = "white", fg = "black", col.axis = "black", col.lab = "black", col.main = "black")
  on.exit(par(op))
  wc.image(wc, which.image = "wc", plot.arrow = TRUE, plot.contour = FALSE,
           color.key = "interval", timelab = "time (days)", periodlab = "period (days)",
           legend.params = list(lab = "wavelet coherence"), main = main, graphics.reset = FALSE)
  abline(h = log2(wc_marks), lty = 2, col = "black", lwd = 1.4)                 # week / month / year guides
  add_xpower_contour(df, wc)             # red-noise 95% CROSS-POWER significance (T&C 1998; kernel)
  usr <- par("usr")
  text(x = usr[1] + 0.01 * (usr[2] - usr[1]), y = log2(wc_marks), labels = names(wc_marks),
       adj = c(0, -0.4), font = 2, cex = 0.9, col = "black")
}

# ---- WaveletComp wrapper --------------------------------------------------------------------
# Analyzed on climate (precip) vs log(response) (log discharge).
# loess.span = 0: no detrending — the analyzed range (4–730 d) sits below any decadal trend,
#   and standardization is then exactly /sd(x), making the ×sd amplitude rescale in band_trace exact.
# my.pair ⇒ Angle = phase.climate − phase.log_response, so climate-leads is +.
# period range spans ~a week → two years so the week / month / year guide lines are all in view
# (see plot_wc). Widening the range leaves the band metrics at P unchanged — band_trace/wave_summary
# read the same ±5%-of-365 scale rows, which sit comfortably inside 4…730 d.
wco <- function(df)
  analyze.coherency(as.data.frame(df[, c("climate", "log_response")]),
                    my.pair = c("climate", "log_response"), loess.span = 0,
                    dt = 1, dj = 1/20, lowerPeriod = 4, upperPeriod = 730,
                    make.pval = FALSE, verbose = FALSE)

# ---- collapse the wc object to one time-resolved series at the P band ------------------------
band_trace <- function(wc, df) {
  bi   <- band_rows(wc)
  sd_c <- sd(df$climate); sd_r <- sd(df$log_response)      # the constants WaveletComp divided out (loess.span=0)
  ang  <- apply(wc$Angle[bi, , drop = FALSE], 2, CMEAN)    # per-time band phase difference
  coi  <- df$time >= EF & df$time <= nrow(df) - EF          # trustworthy (outside COI) mask
  ax   <- colMeans(wc$Ampl.x[bi, , drop = FALSE])          # DRIVER (climate) band amplitude, per time
  ok   <- ax >= PHASE_MIN_AMP * median(ax[coi])            # phase defined only where the driver oscillates at the band
  tibble(time       = df$time, date = df$date,
         phase      = df$phase,
         coherence  = colMeans(wc$Coherence[bi, , drop = FALSE]),
         xpower     = colMeans(wc$Power.xy[bi, , drop = FALSE]),                # shared-oscillation power at band
         amp_ratio  = colMeans(wc$Ampl.y[bi, , drop = FALSE]) /
                      colMeans(wc$Ampl.x[bi, , drop = FALSE]) * (sd_r / sd_c),  # ×sd: undo standardization → physical band-amp ratio
         driver_amp = ax,
         angle      = ifelse(ok, ang, NA_real_),                               # gated: undefined where driver has no band signal
         lag        = ifelse(ok, ang / (2 * pi) * P, NA_real_),                # phase → days (blanked where φ undefined)
         phase_ok   = ok,                                                      # shared determinacy flag (trace + table)
         coi_ok     = coi)
}

# ---- settled pre/post summary with block SEs ------------------------------------------------
# One estimate per P-long (annual) cycle in a COI-buffered window, then value ± dispersion/√n_blocks.
block_stat <- function(t, v, circ = FALSE) {
  b   <- (t - min(t)) %/% P                                # blocks on the FULL window (aligned across metrics)
  bv  <- vapply(split(v, b), function(vi) {                # one estimate per block, dropping gated-out (NA) steps
           vi <- vi[!is.na(vi)]
           if (!length(vi)) NA_real_ else if (circ) CMEAN(vi) else mean(vi)
         }, numeric(1))
  bv  <- bv[!is.na(bv)]                                    # keep only blocks that carried signal
  n   <- length(bv)
  if (circ) {
    R    <- if (n) Mod(mean(exp(1i * bv))) else NA_real_   # resultant length of block phases (1 = locked, 0 = scattered)
    disp <- if (n && R < 1) sqrt(-2 * log(R)) else 0       # circular sd
  } else { R <- NA_real_; disp <- if (n > 1) sd(bv) else NA_real_ }
  list(value = if (!n) NA_real_ else if (circ) CMEAN(bv) else mean(bv),
       se    = if (n > 1) disp / sqrt(n) else NA_real_, n = n, R = R)
}

# Settled windows come from the `phase` column: buffer EF from BOTH the series ends and the
# disturbance. Real phases are pre / disturbance / post (no synthetic "transition"); the
# disturbance may be a single day or a short span (fernow_WS-5), so we buffer from its full extent.
wave_summary <- function(tr) {
  ds <- min(tr$time[tr$phase == "disturbance"])          # first disturbance step
  de <- max(tr$time[tr$phase == "disturbance"])          # last disturbance step (span for WS-5)
  N  <- max(tr$time)
  wins <- list(pre  = tr$time >= EF      & tr$time <= ds - EF,     # ≥ EF from series start AND from disturbance onset
               post = tr$time >= de + EF & tr$time <= N  - EF)     # ≥ EF from disturbance end AND from series end
  one <- function(sel) {
    t  <- tr$time[sel]
    co <- block_stat(t, tr$coherence[sel]); xp <- block_stat(t, tr$xpower[sel])
    ar <- block_stat(t, tr$amp_ratio[sel]); an <- block_stat(t, tr$angle[sel], circ = TRUE)
    # Report φ/lag only where the block phases actually concentrate: need ≥2 signal-bearing blocks
    # (a lone block gives R=1 by construction) AND R ≥ PHASE_MIN_R. Otherwise the phase is scattered
    # (or undefined) and printing a point value would be spurious → NA.
    ph_ok <- !is.na(an$R) && an$n >= 2 && an$R >= PHASE_MIN_R
    phi   <- if (ph_ok) an$value else NA_real_
    phise <- if (ph_ok) an$se    else NA_real_
    tibble(n_blocks  = co$n,                                            # cycles the SEs rest on
           coherence = co$value, coherence_se = co$se,
           xpower    = xp$value, xpower_se    = xp$se,
           amp_ratio = ar$value, amp_ratio_se = ar$se,
           phi       = phi,                    phi_se = phise,           # phase difference φ, radians (circular SE)
           lag       = phi / (2*pi) * P,    lag_se = phise / (2*pi) * P, # same φ in day units
           phase_R   = an$R, phase_n = an$n)                             # phase concentration (0–1) + blocks it rests on
  }
  bind_rows(pre = one(wins$pre), post = one(wins$post), .id = "phase") %>%
    mutate(phase = factor(phase, c("pre", "post")))
}

# ---- model-based residual (whole series, log space) -----------------------------------------
# Predict log(response) from a pre-fit predictor: predicted = gain · climate(t − lag). lag & gain
# are fit on `pre` only (cross-correlation peak + no-intercept OLS), so the residual sits at ~0
# through pre and departs where coupling changes.
fit_pre <- function(df) {
  pre <- df[df$phase == "pre", ]; cl <- pre$climate; rs <- pre$log_response; n <- length(rs)
  Ls  <- 0:round(0.6 * P)                                              # physically plausible lags only (< ~0.6 season)
  cr  <- vapply(Ls, function(L) cor(rs[(L + 1):n], cl[seq_len(n - L)]), numeric(1))  # cor(log(response)(t), climate(t−L))
  L   <- Ls[which.max(cr)]
  x   <- cl[seq_len(n - L)]; y <- rs[(L + 1):n]
  list(lag = L, gain = sum(x * y) / sum(x * x))                        # OLS slope, no intercept
}

resid_trace <- function(df) {
  fp <- fit_pre(df); L <- fp$lag; N <- nrow(df)
  pred <- rep(NA_real_, N); if (L < N) pred[(L + 1):N] <- fp$gain * df$climate[seq_len(N - L)]
  resid <- df$log_response - pred
  out <- tibble(time = df$time, phase = df$phase,
                resid  = resid,
                dresid = c(NA, diff(smooth_ma(resid, 61))))            # smooth THEN diff so d/dt isn't pure noise
  attr(out, "lag") <- L; attr(out, "gain") <- fp$gain; out
}

# ---- shared plotting furniture (band shading matches analysis III) --------------------------
phase_fill <- c(pre = "#9ecae1", disturbance = "grey45", post = "#fdae6b")
bar_fill   <- phase_fill[c("pre", "post")]
phase_bands <- function(ph) ph %>% filter(phase != "disturbance") %>%
  group_by(phase) %>% summarise(xmin = min(time), xmax = max(time), .groups = "drop")

phase_band_layers <- function(d) {
  bands <- phase_bands(d); td <- d$time[d$phase == "disturbance"]
  list(geom_rect(data = bands, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = phase),
                 alpha = 0.30, inherit.aes = FALSE),
       geom_vline(xintercept = td, linetype = 2, colour = "grey70"),
       scale_fill_manual(values = phase_fill, guide = "none"))
}

# alternating calendar-quarter (3-month "season") shading in the plot's x units — a light seasonal
# reference laid over the phase bands so both the disturbance context and the annual rhythm read.
# Needs a `date` column alongside `xcol`.
season_layer <- function(d, xcol = "time") {
  q  <- (as.integer(format(d$date, "%m")) - 1) %/% 3 + 1
  yr <- as.integer(format(d$date, "%Y"))
  r  <- tibble(x = d[[xcol]], qi = yr * 4 + q, q = q) %>%
    group_by(qi) %>% summarise(xmin = min(x), xmax = max(x), odd = first(q) %% 2 == 1, .groups = "drop") %>%
    filter(odd)
  geom_rect(data = r, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
            fill = "grey20", alpha = 0.14, inherit.aes = FALSE)   # every other 3-month season
}

# whole-series residual + its smoothed derivative on a secondary axis, phases shaded AND labelled.
plot_resid <- function(df, title) {
  rt <- resid_trace(df); L <- attr(rt, "lag"); g <- attr(rt, "gain")
  bands <- phase_bands(df)
  ytop <- max(abs(rt$resid), na.rm = TRUE)
  sc   <- ytop / max(abs(rt$dresid), na.rm = TRUE)
  ggplot() +
    phase_band_layers(df) +
    season_layer(df, "time") +
    geom_hline(yintercept = 0, colour = "grey70", linewidth = 0.3) +
    geom_line(data = rt, aes(time, dresid * sc), colour = "#d94801", linewidth = 0.25, na.rm = TRUE) +
    geom_line(data = rt, aes(time, resid),       colour = "#238b45", linewidth = 0.3, na.rm = TRUE) +
    geom_text(data = mutate(bands, x = (xmin + xmax) / 2),
              aes(x = x, y = ytop, label = phase), vjust = 1, size = 3, colour = "grey25") +
    scale_y_continuous(sec.axis = sec_axis(~ . / sc, name = "d(residual)/dt  (smoothed, orange)")) +
    labs(title = title, x = "time (days)", y = "residual = log(response) − predicted  (green)",
         subtitle = sprintf("pre-fit predictor:  log(response)(t) ≈ %.3f · climate(t − %d)", g, L))
}

# time-resolved band metrics; COI-contaminated edges blanked; phases shaded.
metric_labs <- c(coherence = "coherence", xpower = "cross power", amp_ratio = "amp ratio", lag = "app. lag (d)")
# segment id that starts a NEW run at every gap (NA) and — for a circular metric — at every
# ±π branch-cut crossing (adjacent points that jump more than `jump`). geom_line grouped by this
# never connects across a wrap, so the ±P/2 "flip" draws as points on the antiphase rails rather
# than a spurious near-vertical delay change.
seg_id <- function(v, jump = Inf) {
  d   <- abs(c(NA, diff(v)))
  cumsum(is.na(v) | (!is.na(d) & d > jump))
}
plot_band_trace <- function(tr, title) {
  m <- names(metric_labs)
  wide <- tr %>% mutate(across(all_of(m), ~ ifelse(coi_ok, ., NA_real_)))
  long <- bind_rows(lapply(m, function(mm)
    tibble(time = wide$time, metric = mm, value = wide[[mm]],
           seg = seg_id(wide[[mm]], if (mm == "lag") P/2 else Inf)))) %>%   # break lag at wraps
    mutate(metric = factor(metric, m))
  # In the lag facet only: mark 0 and the ±P/2 antiphase rails and name them — so a value pinned
  # near ±P/2 reads as "antiphase (±½ yr)", NOT a real 6-month delay.
  lf   <- factor("lag", m)
  rail <- tibble(metric = lf, y = c(-P/2, P/2))
  zero <- tibble(metric = lf, y = 0)
  note <- tibble(metric = lf, time = max(tr$time), y = P/2, lab = "antiphase (±½ yr)")
  ggplot() +
    phase_band_layers(tr) +
    season_layer(tr, "time") +
    geom_hline(data = rail, aes(yintercept = y), linetype = 2, colour = "grey55", linewidth = 0.3) +
    geom_hline(data = zero, aes(yintercept = y), linetype = 1, colour = "grey55", linewidth = 0.3) +
    geom_text(data = note, aes(time, y, label = lab), hjust = 1, vjust = -0.4, size = 2.6, colour = "grey45") +
    geom_line(data = long, aes(time, value, group = interaction(metric, seg)), linewidth = 0.3, na.rm = TRUE) +
    facet_wrap(~ metric, scales = "free_y", ncol = 1, strip.position = "left",
               labeller = as_labeller(metric_labs)) +
    labs(title = title, x = "time (days)", y = NULL)
}

# reshape a summary tibble to long (id keys × metric × value × se).
metrics_long <- function(sm, keys) {
  m <- names(metric_labs)
  val <- sm %>% select(all_of(c(keys, m))) %>%
    pivot_longer(all_of(m), names_to = "metric", values_to = "value")
  se  <- sm %>% select(all_of(c(keys, paste0(m, "_se")))) %>%
    rename_with(~ sub("_se$", "", .x), all_of(paste0(m, "_se"))) %>%
    pivot_longer(all_of(m), names_to = "metric", values_to = "se")
  left_join(val, se, by = c(keys, "metric")) %>% mutate(metric = factor(metric, m))
}

# pre/post bars per metric, ±1 SE (block SE) so effect size is read against uncertainty.
plot_wave_bars <- function(sm, title) {
  dat <- metrics_long(sm, "phase")
  ggplot(dat, aes(phase, value, fill = phase)) +
    geom_col(width = 0.7) +
    geom_errorbar(aes(ymin = value - se, ymax = value + se), width = 0.25, colour = "black", na.rm = TRUE) +
    facet_wrap(~ metric, scales = "free_y", nrow = 1, labeller = as_labeller(metric_labs)) +
    scale_fill_manual(values = bar_fill, guide = "none") +
    labs(title = title, x = NULL, y = NULL)
}

# the settled pre/post table — same columns every site.
emit_table <- function(sm, caption)
  cat(kable(sm %>% select(phase, n_blocks, coherence, xpower, amp_ratio, phi, lag, phase_R),
            digits = 2, caption = paste(caption,
              "· phase_R = phase concentration among the phase-defined steps (φ/lag → NA if it ever drops < 0.5). A high phase_R is NOT proof of a real annual coupling — read **xpower**: where it is small (chemistry sites, e.g. hbef ≈ 0.03) a large lag is a phase angle at a band the driver barely occupies, not a genuine delay; a strong annual coupling (hjandrews, xpower ≈ 0.4) shows both.")),
      sep = "\n")

# ---- "where does the coupling live" diagnostic ----------------------------------------------
# Time-averaged cross-wavelet power as a function of PERIOD (rows are periods, cols are times in
# the wc matrices → rowMeans collapses time). Shared P–Q power should peak at SHORT (event) periods,
# not at the annual band — the honest picture of an episodic driver.
# `power` = the observed time-mean cross-power; `ref` = the red-noise 95% level for that time-mean
# (analytic AR(1), DOF-inflated for the averaging — see R/wavelet_kernel.R xpower_sig_ref_tavg).
# n_a = the number of time columns rowMeans averages over (the full record).
xpower_spectrum <- function(wc, df, site)
  tibble(site = site, period = wc$Period, power = rowMeans(wc$Power.xy),
         ref  = xpower_sig_ref_tavg(df, wc, n_a = ncol(wc$Power.xy)))

# Annual-harmonic R² (cheap OLS on cos/sin at 365 d): how much of each channel's variance is a
# clean annual cycle. Precip ≈ 0 (aseasonal driver) is the numeric root of the phase degeneracy.
annual_r2 <- function(d) {
  w <- 2 * pi / P; C <- cos(w * d$time); S <- sin(w * d$time)
  r2 <- function(y) summary(lm(y ~ C + S))$r.squared
  tibble(site = d$site[1], precip_annual_r2 = r2(d$climate), logQ_annual_r2 = r2(d$log_response))
}
```

```{r load}
# response (discharge) analyzed in log space → add log_response once at load (prep floored zeros).
sites <- c("hbef_w1", "hjandrews_GSWS10", "fernow_WS-3", "fernow_WS-5")   # MC06 dropped (88% modelled precip driver)
D <- lapply(sites, function(s)
  read_csv(sprintf("%s__series_daily.csv", s), show_col_types = FALSE) %>%
    mutate(phase = factor(phase, c("pre", "disturbance", "post")),
           log_response = log(response)))
names(D) <- sites
# plot labels come from the tracked CSVs themselves (site/disturbance/date carried by the prep) —
# no dependency on any external file, so 04 builds from the committed *__series_daily.csv alone.
# Format: "<site> <disturbance-year> <disturbance>", e.g. "fernow_WS-5 2007 timber_harvest".
label <- setNames(vapply(D, function(d)
  sprintf("%s %s %s", d$site[1],
          format(min(d$date[d$phase == "disturbance"]), "%Y"), d$disturbance[1]),
  character(1)), sites)
```

```{r derive}
# One coherency transform per site → reused by every image, trace and table below (~1 min total).
WC <- lapply(D, wco)
TR <- Map(band_trace, WC, D)
SM <- lapply(TR, wave_summary)
```

## Data provenance (`ms_interp`)

```{r provenance, results='asis'}
prov <- bind_rows(lapply(D, function(d) tibble(
  site = d$site[1],
  climate_interp_pct  = round(100 * mean(d$climate_interp), 1),
  response_interp_pct = round(100 * mean(d$response_interp), 1))))
cat(kable(prov, caption = "Percent of each channel that is MacroSheds gap-fill (modelled), per site (the four analyzed; MC06 dropped for its ~88% modelled precip driver — see intro)."), sep = "\n")
```

---

# Where the shared power lives (and why the annual-band lag is degenerate at some sites)

Before the per-site reads, one diagnostic that reframes everything below. At some sites the
apparent-lag panel at `P≈365` behaves badly — pinning to `±½ yr` and flipping — for a concrete
reason: **where the driver barely oscillates at the annual band**, its phase is near undefined, so
the phase *difference* wanders across `±π`, which the raw `φ→lag` map renders as a `±½ yr` sawtooth
(a circular quantity on a linear axis, not a six-month delay). Un-normalized cross-power vs period
shows *where the shared oscillation actually is* — and, crucially, it **differs by site**.

```{r xpower-spectrum, fig.height=4.2, fig.cap=fig_n("Time-averaged cross-wavelet power vs period (log x), per site, with the red-noise 95% reference (dashed orange). The reference is the analytic AR(1) significance level for a TIME-AVERAGED cross-power (Torrence & Compo 1998, DOF-inflated for the averaging; calibrated against seeded surrogates — tight at event periods, mildly conservative at the annual band). Observed power (green) ABOVE it is shared oscillation beyond what two independent red-noise series would produce. The annual band (grey dashed, 365 d) carries a genuine shared-power peak at hjandrews (dominant, clears the reference) and fernow (alongside a short-period event shoulder), but at hbef the shared power is broadband and event-heavy with only a weak annual bump. Un-normalized cross-power thus separates a real annual coupling from the degenerate case — a distinction the saturated coherence (≈1 everywhere) cannot make.")}
xps <- bind_rows(Map(xpower_spectrum, WC, D, names(WC))) %>% mutate(site = factor(site, sites))
ggplot(xps, aes(period, power)) +
  geom_vline(xintercept = P, linetype = 2, colour = "grey45") +
  geom_line(aes(period, ref), colour = "#d94801", linetype = 2, linewidth = 0.4) +   # red-noise 95%
  geom_line(colour = "#238b45", linewidth = 0.4) +
  facet_wrap(~ site, scales = "free_y") +
  scale_x_log10(breaks = c(7, 30, 91, 365, 730), labels = c("1wk", "1mo", "3mo", "1yr", "2yr")) +
  labs(title = "Where shared power lives: time-averaged cross-power vs period",
       subtitle = "green = observed mean cross-power · dashed orange = red-noise 95% reference",
       x = "period (days, log scale)", y = "mean cross power")
```

```{r annual-r2, results='asis'}
r2tab <- bind_rows(lapply(D, annual_r2))
cat(kable(r2tab, digits = 3,
          caption = "Annual-harmonic R² (OLS on cos/sin at 365 d). Precip's annual cycle is near-zero at hbef/fernow (an aseasonal driver — hence the undefined band phase and ±½ yr wrap there) and only modest (~0.09) at hjandrews; log(discharge) carries the real annual cycle at every site (temperature/snow/ET-driven, not a seasonal precip cycle)."), sep = "\n")
```

So the annual-band lag is trustworthy only where the driver actually oscillates there. At **hbef**
the near-1 coherence coexists with a wrapping, mostly-gated phase — the signature of reading a
nearly-aseasonal driver at a band it does not occupy; its real shared power is broadband. At
**hjandrews** the annual cross-power peak is dominant and the lag sits cleanly at ~`+35 d` (below),
a genuine coupling. In the per-site traces the lag panel is therefore **blanked** wherever the
driver has too little annual-band amplitude to define a phase, and any surviving value is drawn
against `0 / ±½ yr` guides so a rail-hugging cluster reads as *antiphase*, not a six-month delay.

---

# hbef_w1 — chemical addition (pulse)

Chemistry manipulation, not water yield — the seasonal precip→discharge coupling should be a
near-null across the disturbance. A good reference for what the wavelet returns when nothing
hydrologic happened.

```{r hbef-resid, fig.height=3.4, fig.cap=fig_n("hbef_w1 — model residual (green) + smoothed derivative (orange, right axis); phases shaded, dashed = disturbance.")}
plot_resid(D[["hbef_w1"]], paste(label["hbef_w1"], "— model residual & its derivative"))
```

```{r hbef-wc, fig.height=4.6, fig.cap=fig_n("hbef_w1 — wavelet coherence spectrum (time × period). Arrows = phase; pale COI parabola = edge-unreliable. Note how normalized coherence saturates near 1 across most of the annual band.")}
plot_wc(WC[["hbef_w1"]], D[["hbef_w1"]], paste(label["hbef_w1"], "— wavelet coherence"))
```

```{r hbef-trace, fig.height=5.2, fig.cap=fig_n("hbef_w1 — band (P≈365) metrics through time; COI edges blanked, phases shaded.")}
plot_band_trace(TR[["hbef_w1"]], paste(label["hbef_w1"], "— synchrony at the P≈365 band, through time"))
```

```{r hbef-table, results='asis'}
emit_table(SM[["hbef_w1"]], "hbef_w1 — band synchrony, settled pre vs post (n_blocks = annual cycles the SEs rest on)")
```

```{r hbef-bars, fig.height=3, fig.cap=fig_n("hbef_w1 — pre vs post band metrics, ±1 SE (block SE).")}
plot_wave_bars(SM[["hbef_w1"]], paste(label["hbef_w1"], "— band metrics, pre vs post (±1 SE)"))
```

---

# hjandrews_GSWS10 — debris flow (pulse)

A physical channel disturbance that could alter the storm→flow routing. Watch the band amplitude
ratio and apparent lag across 1986. *(~13% of precip is modelled infill.)*

```{r hja-resid, fig.height=3.4, fig.cap=fig_n("hjandrews_GSWS10 — model residual (green) + derivative (orange); phases shaded, dashed = disturbance.")}
plot_resid(D[["hjandrews_GSWS10"]], paste(label["hjandrews_GSWS10"], "— model residual & its derivative"))
```

```{r hja-wc, fig.height=4.6, fig.cap=fig_n("hjandrews_GSWS10 — wavelet coherence spectrum.")}
plot_wc(WC[["hjandrews_GSWS10"]], D[["hjandrews_GSWS10"]], paste(label["hjandrews_GSWS10"], "— wavelet coherence"))
```

```{r hja-trace, fig.height=5.2, fig.cap=fig_n("hjandrews_GSWS10 — band metrics through time.")}
plot_band_trace(TR[["hjandrews_GSWS10"]], paste(label["hjandrews_GSWS10"], "— synchrony at the P≈365 band, through time"))
```

```{r hja-table, results='asis'}
emit_table(SM[["hjandrews_GSWS10"]], "hjandrews_GSWS10 — band synchrony, settled pre vs post")
```

```{r hja-bars, fig.height=3, fig.cap=fig_n("hjandrews_GSWS10 — pre vs post band metrics, ±1 SE.")}
plot_wave_bars(SM[["hjandrews_GSWS10"]], paste(label["hjandrews_GSWS10"], "— band metrics, pre vs post (±1 SE)"))
```

---

# fernow_WS-3 — chemical addition (chronic)

Chronic acidification, open-ended — "post" is a persistently manipulated regime. Chemistry
target, so like hbef expect little hydrologic signature, but the long record gives the wavelet
many annual cycles to average.

```{r fw3-resid, fig.height=3.4, fig.cap=fig_n("fernow_WS-3 — model residual (green) + derivative (orange); phases shaded, dashed = disturbance.")}
plot_resid(D[["fernow_WS-3"]], paste(label["fernow_WS-3"], "— model residual & its derivative"))
```

```{r fw3-wc, fig.height=4.6, fig.cap=fig_n("fernow_WS-3 — wavelet coherence spectrum.")}
plot_wc(WC[["fernow_WS-3"]], D[["fernow_WS-3"]], paste(label["fernow_WS-3"], "— wavelet coherence"))
```

```{r fw3-trace, fig.height=5.2, fig.cap=fig_n("fernow_WS-3 — band metrics through time.")}
plot_band_trace(TR[["fernow_WS-3"]], paste(label["fernow_WS-3"], "— synchrony at the P≈365 band, through time"))
```

```{r fw3-table, results='asis'}
emit_table(SM[["fernow_WS-3"]], "fernow_WS-3 — band synchrony, settled pre vs post")
```

```{r fw3-bars, fig.height=3, fig.cap=fig_n("fernow_WS-3 — pre vs post band metrics, ±1 SE.")}
plot_wave_bars(SM[["fernow_WS-3"]], paste(label["fernow_WS-3"], "— band metrics, pre vs post (±1 SE)"))
```

---

# fernow_WS-5 — timber harvest (pulse)

The clearest water-yield disturbance (2007 clearcut): reduced transpiration raises discharge and
can sharpen the storm response — the site where the seasonal-band amplitude ratio has the most
reason to move.

```{r fw5-resid, fig.height=3.4, fig.cap=fig_n("fernow_WS-5 — model residual (green) + derivative (orange); phases shaded, dashed = disturbance.")}
plot_resid(D[["fernow_WS-5"]], paste(label["fernow_WS-5"], "— model residual & its derivative"))
```

```{r fw5-wc, fig.height=4.6, fig.cap=fig_n("fernow_WS-5 — wavelet coherence spectrum.")}
plot_wc(WC[["fernow_WS-5"]], D[["fernow_WS-5"]], paste(label["fernow_WS-5"], "— wavelet coherence"))
```

```{r fw5-trace, fig.height=5.2, fig.cap=fig_n("fernow_WS-5 — band metrics through time.")}
plot_band_trace(TR[["fernow_WS-5"]], paste(label["fernow_WS-5"], "— synchrony at the P≈365 band, through time"))
```

```{r fw5-table, results='asis'}
emit_table(SM[["fernow_WS-5"]], "fernow_WS-5 — band synchrony, settled pre vs post")
```

```{r fw5-bars, fig.height=3, fig.cap=fig_n("fernow_WS-5 — pre vs post band metrics, ±1 SE.")}
plot_wave_bars(SM[["fernow_WS-5"]], paste(label["fernow_WS-5"], "— band metrics, pre vs post (±1 SE)"))
```

---

# Cross-site summary

Post-minus-pre change in the four band metrics per site, each Δ carrying its propagated SE
(`√(se_pre² + se_post²)`). In analysis II the synthetic scenarios separated cleanly and taught the
lesson that *coherence alone is not enough*; here the question is which of these real sites, if
any, moves a **power/amplitude** channel beyond its (now well-resolved) error.

```{r summary}
sm_all <- bind_rows(SM, .id = "site")
delta <- metrics_long(sm_all, c("site", "phase")) %>%
  pivot_wider(names_from = phase, values_from = c(value, se)) %>%
  mutate(delta    = value_post - value_pre,
         delta_se = sqrt(se_pre^2 + se_post^2),
         site     = factor(site, sites))
```

```{r summary-table, results='asis'}
cat(kable(delta %>% transmute(site, metric = metric_labs[as.character(metric)],
                              pre = value_pre, post = value_post, delta, delta_se),
          digits = 2, caption = "pre → post change in band metrics, all sites (delta_se = propagated ±1 SE)"), sep = "\n")
```

```{r summary-fig, fig.height=4.5, fig.cap=fig_n("Post − pre change per site, faceted by band metric (free y), ±1 propagated SE. The lesson from analysis II carries over: coherence barely moves anywhere (it saturates on real data too), so if anything registers it is in cross-power / amplitude.")}
ggplot(delta, aes(site, delta, fill = site)) +
  geom_col() +
  geom_errorbar(aes(ymin = delta - delta_se, ymax = delta + delta_se), width = 0.25, colour = "black", na.rm = TRUE) +
  geom_hline(yintercept = 0, colour = "grey70") +
  facet_wrap(~ metric, scales = "free_y", labeller = as_labeller(metric_labs)) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(title = "Post − pre change by band metric (free y axes; ±1 SE)", x = NULL, y = "post − pre") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

**Takeaways — the native wavelet view on real synchrony, next to analysis III.** The lesson
analysis II taught on synthetic data reappears **unprompted** on real watersheds: WaveletComp's
*normalized, smoothed* coherence **saturates near 1 across the whole record at every site** (see
the red spectra), so — exactly as in the synthetic Desync case — coherence is not a usable
discriminator here; whatever real change exists must be read from **un-normalized cross-power and
band amplitude**, which the wavelet still resolves with well-populated error bars (decades of
cycles, unlike the synthetic ~2). The tell is site-dependent: where near-1 coherence coexists with
a **wrapping, mostly-gated** annual-band phase (hbef — precip's annual R² ≈ 0), that pairing is the
*signature* of the smoothing artifact, and the real shared power is broadband. Un-normalized
cross-power, by contrast, **separates** that degenerate case from a genuine annual coupling
(hjandrews' dominant annual peak and clean ~35 d lag) — the discrimination the saturated coherence
cannot provide (first figure). Against analysis III: the peak/moments lens frayed on the
episodic precip driver (an unstable "period"), while the wavelet keeps a clean,
time-resolved amplitude/lag read but hands back a **falsely reassuring coherence** — two methods,
two different partial views, each incomplete in its own way. That, not any single site's number,
is the point of running both: the choice of synchrony metric decides what you can and cannot see,
and on real, less-deterministic coupling neither view is sufficient alone.
```
