---
title: "Real-data disturbance–response III — central tendency, variance & peak synchrony"
subtitle: "The peak/moments approach on real MacroSheds precip→discharge — how the simplest synchrony method holds up off the idealised bench (cf. analysis I)"
author: "nic (nic@cascadeclimate.org)"
date: "`r Sys.Date()`"
# Knit → render in place, then copy the self-contained HTML to <repo-root>/docs for GitHub Pages.
knit: |
  (function(input, encoding) {
    out  <- rmarkdown::render(input, encoding = encoding)
    docs <- file.path(dirname(normalizePath(input)), "..", "docs")
    dir.create(docs, showWarnings = FALSE, recursive = TRUE)
    file.copy(out, docs, overwrite = TRUE)
    out
  })
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: darkly
    highlight: breezedark
    fig_width: 10
    fig_caption: true
    self_contained: true
    includes:
      before_body: ../navbar.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dev.args = list(bg = "white"))   # native (light) PNG canvas; white matting added via CSS
suppressMessages({ library(dplyr); library(tidyr); library(readr); library(ggplot2)
                   library(knitr); library(patchwork) })
theme_set(theme_bw(base_size = 11) + theme(legend.position = "bottom"))
P <- 365   # seasonal (annual) period in days — the band this method reads synchrony at; wraps phase lag into [0,P).
fig_n <- local({ i <- 0L; function(x) { i <<- i + 1L; paste0("Figure ", i, ". ", x) } })
```

```{css figure-matting, echo=FALSE}
/* White matting behind every figure: native light plots read cleanly on the darkly page. */
.main-container img,
div.figure img {
  background-color: #ffffff;
  padding: 10px;
  border-radius: 6px;
  box-sizing: border-box;
  max-width: 100%;
  box-shadow: 0 2px 10px rgba(0, 0, 0, 0.45);
}
```

This is **analysis I's method — per-channel moments + peak-derived synchrony — run on real
watersheds** instead of the synthetic ideal. The study is a **methods comparison**: analysis I
(`syn_data/analysis_01`) established, on a clean lagged log-linear system, what this simplest
approach can see (regime amplitude, coarse lag, peak-spacing regularity) and where it is expected
to *fray* (noisy / irregular signals, few cycles). Here we watch it meet **less-deterministic
real synchrony**: `climate` = daily precipitation (the exogenous, undisturbed driver, kept
linear) and `response` = daily discharge (analyzed as `log(response)`), across five screened
disturbances (see `../qualifying_disturbances.csv`, built in `prep_ms_series.Rmd`). Same two
moves as analysis I: **(1)** quantify each channel with central tendency + variance in the
settled `pre` / `post` windows; **(2)** turn those, plus peak spacings, into a synchrony
description across the disturbance. Moments (incl. covariance) and peaks only — no correlation
coefficient, spectra, or wavelets (that is analysis IV's job).

Unlike the synthetic driver, **daily precip is episodic, not a clean sine** and discharge carries
snowmelt/storage lag and long-term trend — so read this
as a stress-test of the peak/moments lens, not a search for a known answer. Where it strains
(peaks fraying on storm-spiky series, an unstable period), that *is* the finding, and analysis IV
takes the same sites into the wavelet view for contrast.

*Error-bar caveat (as in analysis I): the peak-metric SEs are block/sample SEs over the ~decades
of annual cycles in each window — coarser than they look on autocorrelated series; read as
relative guides.*

## Helpers

```{r helpers}
# centered moving average (window w); NA at the ±(w-1)/2 ends — used for peak-finding only.
smooth_ma <- function(x, w = 61) as.numeric(stats::filter(x, rep(1 / w, w)))

# local maxima of a smoothed copy, thinned greedily by height to a minimum separation `gap`.
# Two noise defences: keep only candidates above the window mean, then greedy gap≈0.6P ⇒ ≤1
# peak/season. Real daily series are storm-spiky, so we smooth harder than analysis I (w≈2 months)
# to recover the seasonal peak rather than individual storms.
find_peaks <- function(x, w = 61, gap = round(0.6 * P)) {
  s <- smooth_ma(x, w); n <- length(s)
  interior <- s[2:(n - 1)] > s[1:(n - 2)] & s[2:(n - 1)] >= s[3:n]
  cand <- which(c(FALSE, interior, FALSE))            # NA comparisons drop out of which()
  cand <- cand[s[cand] > mean(s, na.rm = TRUE)]       # peaks lie above central tendency
  cand <- cand[order(s[cand], decreasing = TRUE)]     # tallest first
  keep <- integer(0)
  for (i in cand) if (all(abs(i - keep) >= gap)) keep <- c(keep, i)
  sort(keep)
}

# per-channel moments: central tendency (mean, median) + spread (sd, var, IQR, range).
series_stats <- function(x)
  tibble(mean = mean(x), median = median(x), sd = sd(x),
         var = var(x), IQR = IQR(x), range = diff(range(x)))

# per-response-peak offset behind the preceding climate peak, wrapped into [0,P). Empty if either
# channel has no peak. phase_lag = its mean; the vector also feeds the metric's standard error.
peak_offsets <- function(pc, pr) {
  if (!length(pc) || !length(pr)) return(numeric(0))
  vapply(pr, function(tr) {
    prior <- pc[pc <= tr]
    d <- if (length(prior)) tr - max(prior) else tr - min(pc)
    ((d %% P) + P) %% P
  }, numeric(1))
}
se_mean  <- function(x) if (length(x) > 1) sd(x) / sqrt(length(x)) else NA_real_  # SE of a sample mean

# covariance SE from per-cycle estimates: cut the window into P-long (annual) blocks, cov within
# each, SE = sd(block covs)/√n_blocks. Valid under autocorrelation. Here n_blocks = the ~decades
# of years in each window, so it is far better resolved than the synthetic ~4 cycles.
cov_se <- function(x, y) {
  blk <- (seq_along(x) - 1) %/% P
  cv  <- tapply(seq_along(x), blk, function(i) if (length(i) > 2) cov(x[i], y[i]) else NA_real_)
  se_mean(cv[is.finite(cv)])
}

# synchrony summary for one settled window (climate/response columns). Peaks are found ONCE here
# and every derived quantity — including the mean-type standard errors — reuses them.
# period_resp_sd = spacing irregularity: small ⇒ clean seasonal oscillation, large ⇒ irregular.
sync_one <- function(sub) {
  pc <- find_peaks(sub$climate); pr <- find_peaks(sub$log_response)   # response peaks in log space
  dpr <- diff(pr); off <- peak_offsets(pc, pr)      # spacings / offsets feed both value and SE
  ac <- sd(sub$climate); ar <- sd(sub$log_response) # amp_response = sd(log R); sd computed once
  tibble(n = nrow(sub),
         period_climate     = if (length(pc) > 1) mean(diff(pc)) else NA_real_,
         period_response    = if (length(dpr))    mean(dpr)      else NA_real_,
         period_response_se = se_mean(dpr),          # SE of the mean inter-peak interval
         period_resp_sd     = if (length(dpr) > 1) sd(dpr) else NA_real_,
         amp_climate        = ac,
         amp_response       = ar,
         amp_ratio          = ar / ac,
         phase_lag          = if (length(off)) mean(off) else NA_real_,
         phase_lag_se       = se_mean(off))          # SE of the mean per-peak offset
}

# tables + figure over the settled pre/post windows (disturbance step/span excluded).
stats_by_period <- function(df)
  df %>% filter(phase %in% c("pre", "post")) %>%
    pivot_longer(c(climate, log_response), names_to = "series") %>%
    group_by(phase, series) %>% group_modify(~ series_stats(.x$value)) %>% ungroup() %>%
    mutate(phase = factor(phase, c("pre", "post"))) %>% arrange(phase, series)

sync_by_period <- function(df)
  bind_rows(pre  = sync_one(filter(df, phase == "pre")),
            post = sync_one(filter(df, phase == "post")), .id = "phase") %>%
    mutate(phase = factor(phase, c("pre", "post")))

# contiguous phase spans → shaded background bands; the disturbance (a step or a short span) is
# drawn as a dashed line. climate (mm) and log(response) live on very different scales, so unlike
# analysis I we facet the two channels with free y axes (a shared axis would bury one of them).
phase_fill <- c(pre = "#9ecae1", disturbance = "grey45", post = "#fdae6b")
# alternating calendar-quarter (3-month "season") shading in the plot's x units — a light seasonal
# reference over the phase bands (the annual rhythm; clearest on zoom). Needs a `date` column.
season_layer <- function(d, xcol = "time") {
  q  <- (as.integer(format(d$date, "%m")) - 1) %/% 3 + 1
  yr <- as.integer(format(d$date, "%Y"))
  r  <- tibble(x = d[[xcol]], qi = yr * 4 + q, q = q) %>%
    group_by(qi) %>% summarise(xmin = min(x), xmax = max(x), odd = first(q) %% 2 == 1, .groups = "drop") %>%
    filter(odd)
  geom_rect(data = r, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
            fill = "grey20", alpha = 0.14, inherit.aes = FALSE)
}
plot_series <- function(df, title) {
  bands <- df %>% filter(phase != "disturbance") %>% group_by(phase) %>%
    summarise(xmin = min(time), xmax = max(time), .groups = "drop")
  td <- df$time[df$phase == "disturbance"]
  long <- pivot_longer(df, c(climate, log_response), names_to = "var") %>%
    mutate(var = recode(var, log_response = "log(response) = log discharge",
                             climate = "climate = precip (mm)"))
  ggplot() +
    geom_rect(data = bands, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = phase),
              alpha = 0.35, inherit.aes = FALSE) +
    season_layer(df, "time") +
    geom_vline(xintercept = td, linetype = 2, colour = "grey70") +
    geom_line(data = long, aes(time, value, colour = var), linewidth = 0.2, na.rm = TRUE) +
    facet_wrap(~ var, ncol = 1, scales = "free_y") +
    scale_fill_manual(values = phase_fill, guide = "none") +
    scale_colour_manual(values = c("climate = precip (mm)" = "#3182bd",
                                   "log(response) = log discharge" = "#6a51a3"), guide = "none") +
    labs(title = title, x = "time (days)", y = NULL)
}

# bar companions to the two tables; pre/post reuse the phase-band fills. free scales because
# metrics live on different units.
bar_fill <- c(pre = "#9ecae1", post = "#fdae6b")
plot_moments <- function(stats_x, df, title, errbar = FALSE) {   # central tendency + spread, per channel
  st <- stats_x %>% select(phase, series, mean, median, sd)
  nn <- df %>% filter(phase %in% c("pre", "post")) %>% count(phase, name = "n")   # rows per window
  long <- st %>% left_join(nn, by = "phase") %>%
    mutate(se_mean = sd / sqrt(n)) %>%
    pivot_longer(c(mean, median, sd), names_to = "measure") %>%
    mutate(se = ifelse(measure == "mean", se_mean, NA_real_)) %>%
    select(phase, series, measure, value, se)
  # joint moment: one cov(climate, log_response) per regime → own facet, block (per-cycle) SE.
  cov <- df %>% filter(phase %in% c("pre", "post")) %>% group_by(phase) %>%
    summarise(value = cov(climate, log_response), se = cov_se(climate, log_response), .groups = "drop") %>%
    mutate(phase = factor(phase, c("pre", "post")), series = "clim×log(resp)", measure = "cov")
  long <- bind_rows(long, cov) %>%
    mutate(measure = factor(measure, c("mean", "median", "sd", "cov")))
  ggplot(long, aes(series, value, fill = phase)) +
    geom_col(position = position_dodge(0.8), width = 0.7) +
    { if (errbar) geom_errorbar(aes(ymin = value - se, ymax = value + se),
        position = position_dodge(0.8), width = 0.25, colour = "black", na.rm = TRUE) } +
    facet_wrap(~ measure, scales = "free") +
    scale_fill_manual(values = bar_fill) +
    labs(title = title, x = NULL, y = NULL, fill = NULL) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}

# peak-derived synchrony discriminators; the precomputed sync tibble already carries the mean-type SEs.
plot_sync <- function(sync_x, title) {
  se_long <- sync_x %>%
    transmute(phase, period_response = period_response_se, phase_lag = phase_lag_se) %>%
    pivot_longer(-phase, names_to = "metric", values_to = "se")
  dat <- sync_x %>% select(phase, period_response, amp_ratio, phase_lag, period_resp_sd) %>%
    pivot_longer(-phase, names_to = "metric") %>%
    left_join(se_long, by = c("phase", "metric")) %>%
    mutate(metric = factor(metric, c("period_response", "amp_ratio", "phase_lag", "period_resp_sd")))
  ggplot(dat, aes(phase, value, fill = phase)) +
    geom_col(width = 0.7) +
    geom_errorbar(aes(ymin = value - se, ymax = value + se), width = 0.25, colour = "black", na.rm = TRUE) +
    facet_wrap(~ metric, scales = "free_y", nrow = 1) +
    scale_fill_manual(values = bar_fill, guide = "none") +
    labs(title = title, x = NULL, y = NULL)
}
```

```{r load}
# response (discharge) is analyzed in log space → add log_response once at load. The prep step
# already floored zero-flow days so log() is finite everywhere (see prep_ms_series.Rmd).
sites <- c("hbef_w1", "hjandrews_GSWS10", "fernow_WS-3", "fernow_WS-5")   # MC06 dropped (88% modelled precip driver)
D <- lapply(sites, function(s)
  read_csv(sprintf("%s__series_daily.csv", s), show_col_types = FALSE) %>%
    mutate(phase = factor(phase, c("pre", "disturbance", "post")),
           log_response = log(response)))
names(D) <- sites
# plot labels come from the tracked CSVs themselves (site/disturbance/date carried by the prep) —
# no dependency on any external file, so 03 builds from the committed *__series_daily.csv alone.
# Format: "<site> <disturbance-year> <disturbance>", e.g. "fernow_WS-5 2007 timber_harvest".
label <- setNames(vapply(D, function(d)
  sprintf("%s %s %s", d$site[1],
          format(min(d$date[d$phase == "disturbance"]), "%Y"), d$disturbance[1]),
  character(1)), sites)
```

```{r derive}
# Each site's moment / synchrony summaries are computed ONCE here (find_peaks runs a single time
# per window) and reused by every table, figure, and the cross-site summary below.
STATS <- lapply(D, stats_by_period)
SYNC  <- lapply(D, sync_by_period)
```

## Data provenance (`ms_interp`)

Before reading anything into pre/post differences: **discharge (the `response`) is 100% observed
at every site**, but **precipitation (the `climate` driver) is partly MacroSheds modelled infill**
— ~13% at hjandrews, negligible elsewhere. (`MC06`, whose precip was ~88% modelled, is dropped for
that reason; it remains in the EDA.) Fractions come straight from the `*_interp` flags carried by
`prep_ms_series.Rmd`.

```{r provenance, results='asis'}
prov <- bind_rows(lapply(D, function(d) tibble(
  site = d$site[1],
  climate_interp_pct  = round(100 * mean(d$climate_interp), 1),
  response_interp_pct = round(100 * mean(d$response_interp), 1))))
cat(kable(prov, caption = "Percent of each channel that is MacroSheds gap-fill (modelled), per site"), sep = "\n")
```

---

# hbef_w1 — chemical addition (pulse)

Hubbard Brook W1 received a whole-watershed **calcium (wollastonite) addition** in 1999. The
manipulation targets **chemistry**, not water yield, so the precip→discharge coupling is expected
to be **largely unchanged** — a useful null for this lens (any pre/post move here is climate, not
treatment).

```{r hbef-fig, fig.height=4.4, fig.cap=fig_n("hbef_w1 — precip (climate) & log(discharge) over time, pre/post shaded, dashed = disturbance.")}
plot_series(D[["hbef_w1"]], label["hbef_w1"])
```

```{r hbef-tables, results='asis'}
cat(kable(STATS[["hbef_w1"]], digits = 3, caption = "hbef_w1 — per-channel moments, settled pre/post"), sep = "\n")
cat("\n\n")
cat(kable(SYNC[["hbef_w1"]] %>% select(-ends_with("_se")), digits = 2,
          caption = "hbef_w1 — peak-derived synchrony, pre vs post"), sep = "\n")
```

```{r hbef-bars, fig.height=8, fig.cap=fig_n("hbef_w1 — per-channel moments + joint covariance (top) and peak-derived synchrony metrics (bottom), pre vs post. Error bars = ±1 SE where defined.")}
(plot_moments(STATS[["hbef_w1"]], D[["hbef_w1"]], paste(label["hbef_w1"], "— moments by channel, pre vs post"), errbar = TRUE) /
 plot_sync(SYNC[["hbef_w1"]], paste(label["hbef_w1"], "— synchrony metrics, pre vs post"))) +
  plot_layout(heights = c(2, 1))
```

---

# hjandrews_GSWS10 — debris flow (pulse)

HJ Andrews WS10, a small steep catchment, was scoured by a **1986 debris flow**. A physical
channel disturbance can plausibly shift the storm→flow response; watch whether peak spacing or
amplitude ratio moves beyond climate variability. *(Provenance: ~13% of this site's precip is
modelled infill — see the table above.)*

```{r hja-fig, fig.height=4.4, fig.cap=fig_n("hjandrews_GSWS10 — precip & log(discharge), pre/post shaded, dashed = disturbance.")}
plot_series(D[["hjandrews_GSWS10"]], label["hjandrews_GSWS10"])
```

```{r hja-tables, results='asis'}
cat(kable(STATS[["hjandrews_GSWS10"]], digits = 3, caption = "hjandrews_GSWS10 — per-channel moments, settled pre/post"), sep = "\n")
cat("\n\n")
cat(kable(SYNC[["hjandrews_GSWS10"]] %>% select(-ends_with("_se")), digits = 2,
          caption = "hjandrews_GSWS10 — peak-derived synchrony, pre vs post"), sep = "\n")
```

```{r hja-bars, fig.height=8, fig.cap=fig_n("hjandrews_GSWS10 — moments + covariance (top) and synchrony metrics (bottom), pre vs post; ±1 SE where defined.")}
(plot_moments(STATS[["hjandrews_GSWS10"]], D[["hjandrews_GSWS10"]], paste(label["hjandrews_GSWS10"], "— moments by channel, pre vs post"), errbar = TRUE) /
 plot_sync(SYNC[["hjandrews_GSWS10"]], paste(label["hjandrews_GSWS10"], "— synchrony metrics, pre vs post"))) +
  plot_layout(heights = c(2, 1))
```

---

# fernow_WS-3 — chemical addition (chronic)

Fernow WS3 has received **chronic artificial acidification** (ammonium sulfate) since 1989 — an
*open-ended* treatment, so "post" is a persistently manipulated regime rather than a settled
recovery. Like W1 the target is chemistry; a strong discharge signature is not expected, but the
long records give this method many cycles to average.

```{r fw3-fig, fig.height=4.4, fig.cap=fig_n("fernow_WS-3 — precip & log(discharge), pre/post shaded, dashed = disturbance.")}
plot_series(D[["fernow_WS-3"]], label["fernow_WS-3"])
```

```{r fw3-tables, results='asis'}
cat(kable(STATS[["fernow_WS-3"]], digits = 3, caption = "fernow_WS-3 — per-channel moments, settled pre/post"), sep = "\n")
cat("\n\n")
cat(kable(SYNC[["fernow_WS-3"]] %>% select(-ends_with("_se")), digits = 2,
          caption = "fernow_WS-3 — peak-derived synchrony, pre vs post"), sep = "\n")
```

```{r fw3-bars, fig.height=8, fig.cap=fig_n("fernow_WS-3 — moments + covariance (top) and synchrony metrics (bottom), pre vs post; ±1 SE where defined.")}
(plot_moments(STATS[["fernow_WS-3"]], D[["fernow_WS-3"]], paste(label["fernow_WS-3"], "— moments by channel, pre vs post"), errbar = TRUE) /
 plot_sync(SYNC[["fernow_WS-3"]], paste(label["fernow_WS-3"], "— synchrony metrics, pre vs post"))) +
  plot_layout(heights = c(2, 1))
```

---

# fernow_WS-5 — timber harvest (pulse)

Fernow WS5 was **clearcut in 2007**. Forest removal is the classic water-yield disturbance — reduced
transpiration raises discharge and can sharpen the storm response — so this is the site where the
**precip→discharge lens should have the most to see**.

```{r fw5-fig, fig.height=4.4, fig.cap=fig_n("fernow_WS-5 — precip & log(discharge), pre/post shaded, dashed = disturbance.")}
plot_series(D[["fernow_WS-5"]], label["fernow_WS-5"])
```

```{r fw5-tables, results='asis'}
cat(kable(STATS[["fernow_WS-5"]], digits = 3, caption = "fernow_WS-5 — per-channel moments, settled pre/post"), sep = "\n")
cat("\n\n")
cat(kable(SYNC[["fernow_WS-5"]] %>% select(-ends_with("_se")), digits = 2,
          caption = "fernow_WS-5 — peak-derived synchrony, pre vs post"), sep = "\n")
```

```{r fw5-bars, fig.height=8, fig.cap=fig_n("fernow_WS-5 — moments + covariance (top) and synchrony metrics (bottom), pre vs post; ±1 SE where defined.")}
(plot_moments(STATS[["fernow_WS-5"]], D[["fernow_WS-5"]], paste(label["fernow_WS-5"], "— moments by channel, pre vs post"), errbar = TRUE) /
 plot_sync(SYNC[["fernow_WS-5"]], paste(label["fernow_WS-5"], "— synchrony metrics, pre vs post"))) +
  plot_layout(heights = c(2, 1))
```

---

# Cross-site summary

Post-minus-pre change in the peak-synchrony discriminators across the four sites. Read against
analysis I's synthetic result: there the three scenarios cleanly separated (Shift moved amp
ratio + lag, Desync spiked period-sd, Return sat at zero). On real data the question is different
— **how legibly does this method register anything at all**, and which sites move beyond the
climate-driven pre/post wobble. (`MC06` is dropped — its precip driver is ~88% modelled infill; it
remains in the EDA, where its coverage is now shown.)

```{r summary, fig.height=4}
sync_all <- bind_rows(SYNC, .id = "site")
delta <- sync_all %>%
  select(site, phase, amp_ratio, phase_lag, period_resp_sd) %>%
  pivot_longer(c(amp_ratio, phase_lag, period_resp_sd), names_to = "metric") %>%
  pivot_wider(names_from = phase, values_from = value) %>%
  mutate(delta = post - pre, site = factor(site, sites))
```

```{r summary-table, results='asis'}
cat(kable(delta %>% select(site, metric, pre, post, delta), digits = 2,
          caption = "pre → post change in synchrony discriminators, all sites"), sep = "\n")
```

```{r summary-fig, fig.height=4.5, fig.cap=fig_n("Post − pre change per site, faceted by discriminator (free y). Unlike the synthetic scenarios these do not fall into tidy signatures — the peak/moments lens gives a noisy, partial read on real coupling, motivating the wavelet view in analysis IV.")}
ggplot(delta, aes(site, delta, fill = site)) +
  geom_col() +
  geom_hline(yintercept = 0, colour = "grey70") +
  facet_wrap(~ metric, scales = "free_y",
             labeller = as_labeller(c(amp_ratio = "amp ratio", phase_lag = "phase lag",
                                      period_resp_sd = "response-period sd"))) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(title = "Post − pre change by discriminator (free y axes)", x = NULL, y = "post − pre") +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

**Takeaways — the peak/moments lens on real synchrony.** On the idealised data (analysis I) this
method was decisive; here it is **partial and noisy**. Precip's episodic, non-sinusoidal
character means "peaks" and a stable "period" are already shaky for the driver, the amplitude
ratio no longer maps to a clean β, and real pre/post differences are entangled with decadal
climate variability — so a moves-beyond-error read is hard to defend at most sites. That
limitation is the point: the simplest synchrony description degrades gracefully but loses
resolution off the deterministic bench. **Analysis IV** takes these same five sites into the
Morlet-wavelet view, which resolves coupling jointly in time and frequency and — as the synthetic
Desync case warned — can see (and be fooled in) ways this method cannot.
```
