---
title: "Real-data prep — MacroSheds precip→discharge series with disturbance phases"
subtitle: "Builds the tracked wide per-site CSVs (climate = precip, response = discharge, phase) that analyses 03 & 04 read"
author: "nic (nic@cascadeclimate.org)"
date: "`r Sys.Date()`"
# Knit → render in place, then copy the self-contained HTML to <repo-root>/docs for GitHub Pages.
knit: |
  (function(input, encoding) {
    out  <- rmarkdown::render(input, encoding = encoding)
    docs <- file.path(dirname(normalizePath(input)), "..", "docs")
    dir.create(docs, showWarnings = FALSE, recursive = TRUE)
    file.copy(out, docs, overwrite = TRUE)
    out
  })
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: hide
    theme: darkly
    highlight: breezedark
    fig_width: 10
    fig_caption: true
    self_contained: true
    includes:
      before_body: ../navbar.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      dev.args = list(bg = "white"))
suppressMessages({ library(dplyr); library(tidyr); library(readr); library(ggplot2)
                   library(knitr); library(patchwork) })
theme_set(theme_bw(base_size = 11) + theme(legend.position = "bottom"))
fig_n <- local({ i <- 0L; function(x) { i <<- i + 1L; paste0("Figure ", i, ". ", x) } })
OUT_DIR <- "."   # write the wide CSVs beside this notebook (ms_data/), mirroring syn_data's OUT_DIR
```

```{css figure-matting, echo=FALSE}
.main-container img, div.figure img {
  background-color: #ffffff; padding: 10px; border-radius: 6px;
  box-sizing: border-box; max-width: 100%; box-shadow: 0 2px 10px rgba(0, 0, 0, 0.45);
}
```

This notebook is the **real-data analog of `syn_data/make_synthetic.Rmd`**: it does not analyze
anything, it just assembles the tracked datasets that `analysis_03` / `analysis_04` consume. The
synthetic generator invents a clean lagged log-linear system; here we instead **carve real
MacroSheds watershed records into the same wide `climate / response / phase` shape**, so the two
legs of the study feed identical analysis code (the whole point — a clean methods comparison; see
`CLAUDE.md`).

**Channel mapping** (see `../CLAUDE.md` for the full rationale):

- `climate` = **daily precipitation (mm)** — the exogenous, *undisturbed* driver (the analog of
  the synthetic sine). Kept in **linear** units: precip is non-negative and spiky, its many
  zero-rain days are genuine, and no log is taken on the driver.
- `response` = **daily discharge (L/s)** — the disturbed variable, analyzed in **log space** by
  03/04 (strictly positive, right-skewed, multiplicative). We floor zero-flow days here so
  `log(response)` is always finite (below).
- `phase` = ordered `pre → disturbance → post`, cut from each site's screened disturbance date
  and window in `../qualifying_disturbances.csv`.

Unlike the synthetic ideal, real precip is **episodic (not a clean annual sine)**, discharge
carries **trend / storage / snowmelt lag**, and one site's stream is **intermittent** — exactly
the less-deterministic synchrony 03/04 stress-test the two methods against.

## Inputs & site table

```{r sites}
# The 5 screened disturbances (one per site) with their pre/post windows.
qd <- read_csv("../qualifying_disturbances.csv", show_col_types = FALSE) %>%
  mutate(tag = paste0(domain, "_", site_code))
kable(qd %>% select(tag, disturbance, class, disturbance_start, disturbance_end,
                    window_start, window_end),
      caption = "The 5 qualifying real disturbances (source: ../qualifying_disturbances.csv)")
```

## Build one wide series per site

For each site we (1) intersect its qualifying window with the dates where **both** precip and
discharge exist (this trims `santa_barbara_MC06` — precip starts 2000 — and `hjandrews_GSWS10`
— precip ends 2017), (2) lay both channels on a **continuous daily grid** (both are already
gap-free inside the window, verified — no interpolation), (3) **floor zero-flow days** to half
the 1st-percentile of positive discharge (robust to near-zero traces) so `log(response)` is
finite (flagged in `zero_flow`),
and (4) label `phase` from the disturbance dates. `time` is an integer day index `1…N` — the
unit all the ported peak / wavelet math runs in (`dt = 1` day).

```{r build}
build_site <- function(row) {
  tag <- row$tag
  q <- read_csv(sprintf("../data/%s__discharge_daily.csv",    tag), show_col_types = FALSE)
  p <- read_csv(sprintf("../data/%s__precipitation_daily.csv", tag), show_col_types = FALSE)

  lo <- max(row$window_start, min(q$date), min(p$date))    # common, window-clipped span
  hi <- min(row$window_end,   max(q$date), max(p$date))
  grid <- tibble(date = seq(lo, hi, by = "day"))

  # carry MacroSheds' gap-fill flag (ms_interp == 1) for both channels. Discharge is fully
  # observed at all 5 sites; precipitation is partly modelled-infill at some (hjandrews ~13%,
  # MC06 ~88%) — surfaced here and flagged per row so the analyses can caveat it.
  qv <- q %>% filter(date >= lo, date <= hi) %>% transmute(date, discharge = val, q_interp = ms_interp == 1)
  pv <- p %>% filter(date >= lo, date <= hi) %>% transmute(date, precip    = val, p_interp = ms_interp == 1)
  d  <- grid %>% left_join(qv, by = "date") %>% left_join(pv, by = "date")

  # Corrupt-precip guard. MacroSheds' Santa Barbara (MC06) precip carries impossible gap-fill
  # spikes (~6.15e6 mm/d, all ms_interp == 1); real daily precip here is < 170 mm/d. Flag values
  # > 500 mm/d, blank them, and linearly interpolate on the daily grid so the driver stays
  # continuous (WaveletComp needs a gap-free series). Only MC06 is affected; harmless elsewhere.
  d$climate_corrupt <- !is.na(d$precip) & d$precip > 500
  d$precip[d$climate_corrupt] <- NA_real_
  if (anyNA(d$precip)) d$precip <- approx(seq_len(nrow(d)), d$precip, seq_len(nrow(d)), rule = 2)$y

  # log floor = half the 1st-percentile of positive flow. Robust to near-zero trace values (e.g.
  # hbef has positive readings ~3e-5 L/s); flooring at the raw minimum would put no-flow days at an
  # absurd log outlier (~ -11) that dominates the variance. The 1st-pct floor keeps them "very low".
  eps <- 0.5 * as.numeric(quantile(d$discharge[d$discharge > 0], 0.01, na.rm = TRUE))
  ds <- row$disturbance_start; de <- row$disturbance_end
  d %>% transmute(
    time     = row_number(),
    date,
    climate  = precip,                                   # linear driver (mm)
    response = pmax(discharge, eps),                     # floored discharge (L/s), strictly positive
    zero_flow = discharge <= 0,                          # days that were floored (intermittent flow)
    climate_interp  = p_interp,                          # precip day is MacroSheds gap-fill (modelled)
    climate_corrupt = climate_corrupt,                   # precip day was an impossible spike, blanked+interpolated (MC06 only)
    response_interp = q_interp,                          # discharge day is gap-fill (0% at every site)
    phase = factor(ifelse(date < ds, "pre",
                   ifelse(date <= de, "disturbance", "post")),
                   levels = c("pre", "disturbance", "post")),
    site  = tag,
    disturbance = row$disturbance,                       # carried so 03/04 need no external file
    class = row$class)
}

series <- lapply(seq_len(nrow(qd)), function(i) build_site(qd[i, ]))
names(series) <- qd$tag
```

## Write the tracked CSVs & summary

```{r write}
paths <- vapply(names(series), function(tag) {
  f <- file.path(OUT_DIR, paste0(tag, "__series_daily.csv"))
  # response rounded to 6 dp (not 4): some sites have sub-1e-4 flows that 4 dp would zero out → -Inf in log.
  write_csv(series[[tag]] %>% mutate(climate = round(climate, 4), response = round(response, 6)), f)
  f
}, character(1))

summ <- bind_rows(lapply(series, function(s) tibble(
  site = s$site[1], class = s$class[1], n_days = nrow(s),
  start = min(s$date), end = max(s$date),
  pre_days  = sum(s$phase == "pre"),
  post_days = sum(s$phase == "post"),
  zero_flow_days = sum(s$zero_flow),
  climate_corrupt_days = sum(s$climate_corrupt),                  # impossible precip spikes blanked+interpolated (MC06 only)
  climate_interp_pct  = round(100 * mean(s$climate_interp), 1),   # % of precip days that are modelled infill
  response_interp_pct = round(100 * mean(s$response_interp), 1),  # % of discharge days infilled (≈0)
  med_climate = median(s$climate), med_response = median(s$response))))
kable(summ, digits = 2, caption = "Wide series written to ms_data/ (one row per site)")
```

**Data-provenance note (`ms_interp`).** Discharge is **fully observed** (0% gap-filled) at every
site — the `log(response)` channel is real throughout. Precipitation, however, is partly
**MacroSheds modelled infill**: negligible at `hbef_w1` / `fernow_WS-3` / `fernow_WS-5`, ~13% at
`hjandrews_GSWS10`, and **~88% at `santa_barbara_MC06`** (its precip driver is almost entirely
gap-filled — treat any MC06 precip→discharge coupling as coupling to a *modelled* driver, not an
observed one). Analyses 03/04 repeat this caveat where it matters.

## Sanity panels

Climate (precip, top) and `log(response)` (log discharge, bottom) per site, phases shaded, the
disturbance drawn as a dashed line. This is the same view `analysis_03` opens each site section
with — here purely to confirm the join, phase cut, and log-floor look right.

```{r panels, fig.height=13, fig.cap=fig_n("Per-site assembled series: precip (blue) and log(discharge) (purple), pre/post shaded, dashed = disturbance date. Note the episodic precip driver and, at MC06, the intermittent-flow floor in log space.")}
phase_fill <- c(pre = "#9ecae1", disturbance = "grey45", post = "#fdae6b")
one_panel <- function(s) {
  bands <- s %>% filter(phase != "disturbance") %>% group_by(phase) %>%
    summarise(xmin = min(time), xmax = max(time), .groups = "drop")
  td <- s$time[s$phase == "disturbance"]
  long <- s %>% transmute(time, climate, log_response = log(response)) %>%
    pivot_longer(c(climate, log_response), names_to = "var")
  ggplot() +
    geom_rect(data = bands, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = phase),
              alpha = 0.35, inherit.aes = FALSE) +
    geom_vline(xintercept = td, linetype = 2, colour = "grey60") +
    geom_line(data = long, aes(time, value, colour = var), linewidth = 0.2) +
    facet_wrap(~ var, ncol = 1, scales = "free_y") +
    scale_fill_manual(values = phase_fill, guide = "none") +
    scale_colour_manual(values = c(climate = "#3182bd", log_response = "#6a51a3"), guide = "none") +
    labs(title = s$site[1], x = "time (days)", y = NULL)
}
wrap_plots(lapply(series, one_panel), ncol = 1)
```

The five `*__series_daily.csv` files are the **only tracked interface** between this prep and the
two analyses — exactly as `synthetic_*.csv` are for `syn_data`. Regenerate them by knitting this
notebook (it reads `../data/` + `../qualifying_disturbances.csv`, which require the local
MacroSheds pull but are not needed to *run* 03/04).
```
